---
title: "SVAR Analysis of the Italian Electricity Market"
subtitle: "Identification by Sign Restrictions (Kilian & Lütkepohl, 2017, Ch. 13)"
author: "Alessandro Macchi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")

source("00_config.R")
source("01_load_data.R")
source("02_transform.R")
source("03_sanity_checks.R")
source("04_var_model.R")
source("05_sign_restrictions.R")
source("06_bootstrap.R")
source("07_fevd.R")

dir.create(config$paths$output_dir, showWarnings = FALSE, recursive = TRUE)
```

# Overview

This report estimates a reduced-form VAR for the Italian electricity market
and identifies structural shocks via **sign restrictions**, following Kilian
& Lütkepohl (2017, Ch. 13): Haar-uniform random rotations of the Cholesky
factor (via QR decomposition), an admissible-set search over the configured
sign restrictions, and Fry & Pagan (2011) median-target (MT) selection for
the reported impulse responses. Confidence bands come from a **frequentist**
residual bootstrap (Inoue & Kilian, 2013) that re-runs the entire
identification search inside every replication.

Every modeling choice below (variables, transformations, lags, restrictions,
bootstrap settings) is read from `00_config.R` -- that is the only file that
needs editing to change the analysis.

## Endogenous variables

```{r vars-table}
var_table <- purrr::map_dfr(config$variables, ~ tibble::tibble(
  id = .x$id, file = .x$file, label = .x$label,
  transform = .x$transform, role = .x$role
))
knitr::kable(var_table, caption = "Variables configured in 00_config.R")
```

# Data loading

Each raw CSV is parsed with its own delimiter, date format, and decimal
mark (per-variable settings in `00_config.R`), then aligned onto one common
monthly grid so any month missing from a source file surfaces as an
explicit `NA` instead of silently shifting the other series out of
alignment.

```{r load-data}
raw_data <- load_all_series(config)
cat(sprintf("Loaded %d monthly observations, %s to %s.\n",
            nrow(raw_data), min(raw_data$date), max(raw_data$date)))
knitr::kable(head(raw_data, 6), digits = 2, caption = "Raw data (first 6 months)")
```

```{r plot-fns}
## One look for every figure in this report: colours from grDevices'
## Okabe-Ito colourblind-safe palette, no plot box, horizontal tick labels,
## plain left-aligned titles, and a light horizontal grid behind the data.
## `theme_par(...)` takes the per-figure par settings (mfrow, mar) too, so
## each plotting chunk sets its layout and the style in one call.
pal <- grDevices::palette.colors(NULL, "Okabe-Ito")

theme_par <- function(...) {
  par(bty = "n", las = 1, tcl = -0.25, mgp = c(2.4, 0.5, 0), adj = 0,
      font.main = 1, cex.main = 1.05, col.main = "grey15",
      col.axis = "grey35", col.lab = "grey30", fg = "grey65", ...)
}

grid_h <- function() grid(nx = NA, ny = NULL, col = "grey92", lty = 1, lwd = 0.7)

#' Draw one plot into `plots_dir/<variable id>/<name>.png` instead of the
#' report -- the report shows the combined grids, the per-variable files are
#' for use outside it.
save_plot <- function(id, name, draw) {
  dir <- file.path(config$paths$plots_dir, id)
  dir.create(dir, showWarnings = FALSE, recursive = TRUE)
  grDevices::png(file.path(dir, paste0(name, ".png")),
                 width = 7, height = 4.5, units = "in", res = 150)
  on.exit(grDevices::dev.off())
  theme_par(mar = c(3.5, 4, 3, 1))
  draw()
}

plot_series <- function(v) {
  plot(raw_data$date, raw_data[[v$id]], type = "l", col = pal[["blue"]], lwd = 1.5,
       xlab = "", ylab = "", main = v$label, panel.first = grid_h())
}

## NA-dropping matches the unit-root tests in 03_sanity_checks.R: the common
## monthly grid pads series that start late with explicit NAs.
plot_acf  <- function(v) stats::acf(stats::na.omit(raw_data[[v$id]]), lag.max = 36,
                                    main = paste("ACF:", v$label),
                                    col = pal[["blue"]], lwd = 1.5,
                                    ci.col = pal[["vermillion"]], panel.first = grid_h())
plot_pacf <- function(v) stats::pacf(stats::na.omit(raw_data[[v$id]]), lag.max = 36,
                                     main = paste("PACF:", v$label),
                                     col = pal[["blue"]], lwd = 1.5,
                                     ci.col = pal[["vermillion"]], panel.first = grid_h())
```

```{r plot-raw, fig.width=9, fig.height=8}
theme_par(mfrow = c(ceiling(length(config$variables) / 2), 2), mar = c(3, 4, 3, 1))
for (v in config$variables) plot_series(v)

for (v in config$variables) save_plot(v$id, "timeseries", function() plot_series(v))
```

## Autocorrelation

Sample ACF and PACF of each raw series, over 36 monthly lags. The slow decay
in the level series is what the transformations in `00_config.R` address;
spikes at lags 12, 24 and 36 (clearest in electricity consumption) are the
calendar seasonality the monthly dummies absorb.

```{r plot-acf, fig.width=9, fig.height=14}
## Wider top/bottom margins than the raw-series grid: with 12 panels the
## default c(3, 4, 2, 1) leaves too little room and each panel title lands on
## the "Lag" axis of the panel above it.
theme_par(mfrow = c(length(config$variables), 2), mar = c(4, 4, 3, 1))
for (v in config$variables) {
  plot_acf(v)
  plot_pacf(v)
}

for (v in config$variables) {
  save_plot(v$id, "acf", function() plot_acf(v))
  save_plot(v$id, "pacf", function() plot_pacf(v))
}
```

# Transformations

Each variable's `transform` field (`level` / `log` / `diff` / `logdiff`) is
applied generically -- switching a transformation is a one-line change in
`00_config.R` and never touches this code.

```{r transform}
model_data <- transform_data(raw_data, config$variables)
knitr::kable(head(model_data, 6), digits = 4, caption = "Transformed data (first 6 months)")
```

# VAR estimation

Variables are split into endogenous/exogenous automatically from each
variable's `role` in config. Eleven monthly seasonal dummies (Feb-Dec; Jan
is the reference month) are always appended to the exogenous block to
control for calendar seasonality. The lag order follows
`config$lag_selection` (`fixed` or `auto` via `vars::VARselect()`).

```{r var-estimation}
var_out <- estimate_var(model_data, config)
cat(sprintf("Lag order p = %d (method: %s). Estimation sample: %d observations.\n",
            var_out$p, config$lag_selection$method, nrow(var_out$endo) - var_out$p))
```

# Diagnostics

Per the thesis's model-validation requirements: unit-root tests (ADF,
KPSS), a Johansen cointegration test, VAR residual diagnostics
(autocorrelation, normality), and companion-matrix stability. One compact
table per check -- read the statistics against the reported critical
values / p-values directly, no automatic interpretation.

```{r unit-root-level}
knitr::kable(unit_root_table(raw_data, config$variables, space = "level"), digits = 3,
             caption = "Unit-root tests, level (ADF H0: unit root; KPSS H0: stationary; p < 0.05 rejects H0)")
```

```{r unit-root-transform}
knitr::kable(unit_root_table(raw_data, config$variables, space = "transform"), digits = 3,
             caption = "Unit-root tests, after transform (ADF H0: unit root; KPSS H0: stationary; p < 0.05 rejects H0)")
```

## Deterministic components and unit roots

```{r stationarity-table}
stationarity <- stationarity_table(raw_data, config)
knitr::kable(stationarity, digits = 3, caption = stationarity_note(stationarity))
```

```{r za-table}
knitr::kable(za_table(raw_data, config), digits = 3,
             caption = "Zivot-Andrews, endogenous level shift (H0: unit root with no break)")
```

```{r johansen-test}
knitr::kable(johansen_test(raw_data, config$variables, var_out$p), digits = 3,
             caption = "Johansen trace test (H0: cointegration rank <= r), 5% critical value")
```

```{r residual-diagnostics}
knitr::kable(residual_diagnostics(var_out$fit, var_out$p), digits = 3,
             caption = "VAR residual diagnostics")
```

```{r stability-table}
knitr::kable(stability_table(var_out$fit), digits = 3,
             caption = "Companion-matrix root moduli (sorted descending)")
```

# Identification

Structural shocks are identified via sign restrictions read generically
from `config$shocks` -- adding a shock or a restriction requires no changes
to `05_sign_restrictions.R`. For each of `config$identification$n_draws`
Haar-uniform random rotations $Q$, the structural impact matrix
$B_0 = P Q$ is checked against every configured shock's sign restrictions;
admissible draws are kept, and the **Fry-Pagan median-target** draw (the
single admissible model closest to the pointwise median IRF) is reported,
rather than the pointwise median itself.

```{r identification}
ident <- identify_sign_restrictions(var_out$fit, config)

cat(sprintf(
  "Acceptance rate: %.2f%% (%d / %d rotations admissible).\n",
  100 * ident$draws$acceptance_rate, ident$draws$n_accepted, ident$draws$n_draws
))
```

The acceptance rate is a diagnostic of how informative the restrictions
are: a low rate means the restrictions rule out most of the rotation
space (informative, but a non-refutability check worth keeping in mind);
a rate near 100% means the restrictions barely bind.

```{r restrictions-table}
restr_table <- purrr::map_dfr(config$shocks, function(s) {
  purrr::imap_dfr(s$restrictions, function(r, var_id) {
    tibble::tibble(shock = s$name, variable = var_id, sign = r$sign,
                    horizons = paste(range(r$horizons), collapse = ":"))
  })
})
knitr::kable(restr_table, caption = "Configured sign restrictions")
```

# Bootstrap

Frequentist confidence bands (Inoue & Kilian, 2013): the reduced-form VAR
residuals are resampled, the sample is rebuilt recursively, the VAR is
re-estimated, and the **entire identification search above is re-run**
inside every replication -- not just a coefficient resample.

```{r bootstrap}
boot_out <- run_bootstrap(var_out$fit, var_out$endo, var_out$p, config)

cat(sprintf(
  "Bootstrap: %d replications, %d succeeded, %d found no admissible rotation.\n",
  boot_out$n_boot, boot_out$n_ok, boot_out$n_failed
))
cat(sprintf(
  "Mean acceptance rate across successful replications: %.2f%%.\n",
  100 * mean(boot_out$boot_acceptance_rates)
))
```

```{r boot-accept-hist, fig.width=7, fig.height=4.5}
theme_par(mar = c(4, 4, 3, 1))
h <- hist(100 * boot_out$boot_acceptance_rates, breaks = 15, plot = FALSE)
## Drawn twice so the grid sits behind the bars: plot.histogram() has no
## panel.first.
plot(h, col = NA, border = NA, main = "Bootstrap replications: acceptance rate distribution",
     xlab = "Acceptance rate (%)", ylab = "Replications")
grid_h()
plot(h, col = pal[["skyblue"]], border = "white", add = TRUE)
```

# Results

Median-target impulse responses with `r sprintf("%.0f%%", 100 * config$bootstrap$conf_level)`
bootstrap confidence bands. Every endogenous variable enters the VAR in logs, so
the responses are percent deviations from baseline to a one-standard-deviation
structural shock (log points x 100; exactly, 100(e^theta - 1)).

```{r irf-plot-fn}
plot_irf_grid <- function(mt_irf, bands, var_names, shock_names, var_labels) {
  n_shocks <- length(shock_names)
  n_vars <- length(var_names)
  h <- seq_len(dim(mt_irf)[3]) - 1

  old_par <- theme_par(mfrow = c(n_vars, n_shocks), mar = c(3, 4, 3, 1))
  on.exit(par(old_par))

  for (v in seq_len(n_vars)) {
    for (s in seq_len(n_shocks)) {
      ## All endogenous variables enter the VAR in logs, so the responses are
      ## log points; x100 reads them as percent deviations from baseline.
      point <- 100 * mt_irf[v, s, ]
      lo <- 100 * bands$lower[v, s, ]
      hi <- 100 * bands$upper[v, s, ]

      plot(h, point, type = "n", ylim = range(c(point, lo, hi), na.rm = TRUE),
           xlab = if (v == n_vars) "Horizon (months)" else "",
           ylab = paste0(var_labels[v], " (%)"),
           main = if (v == 1) gsub("_", " ", shock_names[s]) else "",
           panel.first = grid_h())
      polygon(c(h, rev(h)), c(lo, rev(hi)),
              col = grDevices::adjustcolor(pal[["blue"]], alpha.f = 0.18), border = NA)
      abline(h = 0, lty = 2, col = "grey55")
      lines(h, point, col = pal[["blue"]], lwd = 2)
    }
  }
}

## config labels carry the RAW series' units ("PUN (EUR/MWh)"), which are wrong
## on an IRF axis: the responses are percent deviations, not levels. Strip the
## trailing unit parenthesis -- plot_irf_grid() appends "(%)" instead. Only the
## IRF grid uses these; the raw-series and ACF panels read v$label directly.
var_labels <- sub("\\s*\\(.*\\)$", "", purrr::map_chr(config$variables, "label"))
names(var_labels) <- purrr::map_chr(config$variables, "id")
```

```{r irf-plot, fig.width=9, fig.height=10}
plot_irf_grid(ident$mt$mt_irf, boot_out$bands, ident$var_names, ident$shock_names,
              var_labels[ident$var_names])
```

## Variance decomposition

Share of each variable's h-step forecast error variance attributable to the
identified shocks, for the same Fry-Pagan median-target model as the impulse
responses above. Horizons are months after impact, so h = 0 is the impact
period. Identification is **partial** -- only the shocks in `config$shocks`
are pinned down -- so the denominator is the full reduced-form forecast error
variance implied by $\Sigma_u$, not the sum over the identified columns of
$B_0$. The identified shares consequently sum to less than one, and the
`unidentified` column is the remainder left to the shocks the sign
restrictions do not identify. It is reported as-is: rescaling the shares to
sum to one would claim an identification this design does not deliver.

Brackets are `r sprintf("%.0f%%", 100 * config$bootstrap$conf_level)` pointwise
percentile bands over the same pooled bootstrap draws as the IRF bands -- the
FEVD is computed inside each replication, so no second bootstrap pass is run.

```{r fevd-table}
fevd_mt <- fevd_median_target(var_out$fit, ident, config$identification$horizon)
fevd_names <- c(gsub("_", " ", ident$shock_names), "unidentified")

fevd_tbl <- purrr::imap_dfr(config$fevd$horizons, function(h, k) {
  tbl <- tibble::tibble(horizon = h, variable = unname(var_labels[ident$var_names]))
  for (b in seq_along(fevd_names)) {
    tbl[[fevd_names[b]]] <- sprintf("%.2f [%.2f, %.2f]", fevd_mt[, b, h + 1],
                                     boot_out$fevd_bands$lower[, b, k],
                                     boot_out$fevd_bands$upper[, b, k])
  }
  tbl
})
knitr::kable(fevd_tbl, caption = paste0(
  "Forecast error variance decomposition, median-target model, with ",
  sprintf("%.0f%%", 100 * config$bootstrap$conf_level), " bootstrap bands"))
```

```{r fevd-plot, fig.width=9, fig.height=7}
#' Stacked-area FEVD, one panel per variable: the identified shocks' shares
#' stacked from the bottom and the unidentified remainder filling the gap up
#' to one, so how much of each variable the restrictions actually explain is
#' the visible grey band rather than something the reader has to subtract.
plot_fevd_grid <- function(fevd, band_names, var_labels) {
  h <- seq_len(dim(fevd)[3]) - 1
  n_vars <- dim(fevd)[1]
  cols <- c(unname(pal[c("blue", "orange", "bluishgreen", "reddishpurple")])[
    seq_len(length(band_names) - 1)], "grey85")

  old_par <- theme_par(mfrow = c(ceiling(n_vars / 2), 2), mar = c(3.5, 4, 3, 1))
  on.exit(par(old_par))

  for (v in seq_len(n_vars)) {
    cum <- rbind(0, apply(fevd[v, , ], 2, cumsum))
    plot(h, cum[nrow(cum), ], type = "n", ylim = c(0, 1),
         xlab = if (v > n_vars - 2) "Horizon (months)" else "",
         ylab = "Share of forecast error variance",
         main = var_labels[v], panel.first = grid_h())
    for (b in seq_along(band_names)) {
      polygon(c(h, rev(h)), c(cum[b, ], rev(cum[b + 1, ])), col = cols[b], border = NA)
    }
    if (v == 1) {
      legend("bottomright", legend = band_names, fill = cols, border = NA,
             bty = "o", box.col = NA, cex = 0.85,
             bg = grDevices::adjustcolor("white", alpha.f = 0.85))
    }
  }
}

plot_fevd_grid(fevd_mt, fevd_names, var_labels[ident$var_names])
```

## Saved output

```{r save-output}
saveRDS(list(config = config, var_out = var_out, ident = ident, boot_out = boot_out,
             fevd_mt = fevd_mt),
        file.path(config$paths$output_dir, "svar_results.rds"))
cat("Saved results to", file.path(config$paths$output_dir, "svar_results.rds"), "\n")
```
