---
title: "SVAR Analysis of the Italian Electricity Market"
subtitle: "Identification by Sign Restrictions (Kilian & Lütkepohl, 2017, Ch. 13)"
author: "Alessandro Macchi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")

source("00_config.R")
source("01_load_data.R")
source("02_transform.R")
source("03_sanity_checks.R")
source("04_var_model.R")
source("05_sign_restrictions.R")
source("06_bootstrap.R")
source("07_fevd.R")
source("08_historical_decomposition.R")
source("09_conditional_forecast.R")

## Month names in English whatever the system locale: the conditional-forecast
## tables and date axes print them.
invisible(Sys.setlocale("LC_TIME", "C"))

dir.create(config$paths$output_dir, showWarnings = FALSE, recursive = TRUE)
```

# Overview

This report estimates a reduced-form VAR for the Italian electricity market
and identifies structural shocks via **sign restrictions**, following Kilian
& Lütkepohl (2017, Ch. 13): Haar-uniform random rotations of the Cholesky
factor (via QR decomposition), an admissible-set search over the configured
sign restrictions, and Fry & Pagan (2011) median-target (MT) selection for
the reported impulse responses. Confidence bands come from a **frequentist**
residual bootstrap (Inoue & Kilian, 2013) that re-runs the entire
identification search inside every replication.

Every modeling choice below (variables, transformations, lags, restrictions,
bootstrap settings) is read from `00_config.R` -- that is the only file that
needs editing to change the analysis.

## Endogenous variables

```{r vars-table}
var_table <- purrr::map_dfr(config$variables, ~ tibble::tibble(
  id = .x$id, file = .x$file, label = .x$label,
  transform = .x$transform, role = .x$role
))
knitr::kable(var_table, caption = "Variables configured in 00_config.R")
```

# Data loading

Each raw CSV is parsed with its own delimiter, date format, and decimal
mark (per-variable settings in `00_config.R`), then aligned onto one common
monthly grid so any month missing from a source file surfaces as an
explicit `NA` instead of silently shifting the other series out of
alignment.

```{r load-data}
raw_data <- load_all_series(config)
cat(sprintf("Loaded %d monthly observations, %s to %s.\n",
            nrow(raw_data), min(raw_data$date), max(raw_data$date)))
knitr::kable(head(raw_data, 6), digits = 2, caption = "Raw data (first 6 months)")
```

```{r plot-fns}
## One look for every figure in this report: colours from grDevices'
## Okabe-Ito colourblind-safe palette, no plot box, horizontal tick labels,
## plain left-aligned titles, and a light horizontal grid behind the data.
## `theme_par(...)` takes the per-figure par settings (mfrow, mar) too, so
## each plotting chunk sets its layout and the style in one call.
pal <- grDevices::palette.colors(NULL, "Okabe-Ito")

theme_par <- function(...) {
  par(bty = "n", las = 1, tcl = -0.25, mgp = c(2.4, 0.5, 0), adj = 0,
      font.main = 1, cex.main = 1.05, col.main = "grey15",
      col.axis = "grey35", col.lab = "grey30", fg = "grey65", ...)
}

grid_h <- function() grid(nx = NA, ny = NULL, col = "grey92", lty = 1, lwd = 0.7)

#' Legend in the figure's top outer margin. The FEVD and historical
#' decomposition panels are filled edge to edge, so a legend placed inside any
#' one of them sits on the data. Pair with `oma = c(0, 0, 3, 0)` in the
#' theme_par() call so the outer margin exists to draw into.
top_legend <- function(...) {
  legend(grconvertX(0.5, "ndc", "user"), grconvertY(1, "ndc", "user"),
         xjust = 0.5, yjust = 1, xpd = NA,
         border = NA, bty = "n", cex = 0.85, ...)
}

#' Draw one plot into `plots_dir/<variable id>/<name>.png` instead of the
#' report -- the report shows the combined grids, the per-variable files are
#' for use outside it.
save_plot <- function(id, name, draw) {
  dir <- file.path(config$paths$plots_dir, id)
  dir.create(dir, showWarnings = FALSE, recursive = TRUE)
  grDevices::png(file.path(dir, paste0(name, ".png")),
                 width = 7, height = 4.5, units = "in", res = 150)
  on.exit(grDevices::dev.off())
  theme_par(mar = c(3.5, 4, 3, 1))
  draw()
}

plot_series <- function(v) {
  plot(raw_data$date, raw_data[[v$id]], type = "l", col = pal[["blue"]], lwd = 1.5,
       xlab = "", ylab = "", main = v$label, panel.first = grid_h())
}

## NA-dropping matches the unit-root tests in 03_sanity_checks.R: the common
## monthly grid pads series that start late with explicit NAs.
plot_acf  <- function(v) stats::acf(stats::na.omit(raw_data[[v$id]]), lag.max = 36,
                                    main = paste("ACF:", v$label),
                                    col = pal[["blue"]], lwd = 1.5,
                                    ci.col = pal[["vermillion"]], panel.first = grid_h())
plot_pacf <- function(v) stats::pacf(stats::na.omit(raw_data[[v$id]]), lag.max = 36,
                                     main = paste("PACF:", v$label),
                                     col = pal[["blue"]], lwd = 1.5,
                                     ci.col = pal[["vermillion"]], panel.first = grid_h())
```

```{r plot-raw, fig.width=9, fig.height=8}
theme_par(mfrow = c(ceiling(length(config$variables) / 2), 2), mar = c(3, 4, 3, 1))
for (v in config$variables) plot_series(v)

for (v in config$variables) save_plot(v$id, "timeseries", function() plot_series(v))
```

## Autocorrelation

Sample ACF and PACF of each raw series, over 36 monthly lags. The slow decay
in the level series is what the transformations in `00_config.R` address;
spikes at lags 12, 24 and 36 (clearest in electricity consumption) are the
calendar seasonality the monthly dummies absorb.

```{r plot-acf, fig.width=9, fig.height=14}
## Wider top/bottom margins than the raw-series grid: with 12 panels the
## default c(3, 4, 2, 1) leaves too little room and each panel title lands on
## the "Lag" axis of the panel above it.
theme_par(mfrow = c(length(config$variables), 2), mar = c(4, 4, 3, 1))
for (v in config$variables) {
  plot_acf(v)
  plot_pacf(v)
}

for (v in config$variables) {
  save_plot(v$id, "acf", function() plot_acf(v))
  save_plot(v$id, "pacf", function() plot_pacf(v))
}
```

# Transformations

Each variable's `transform` field (`level` / `log` / `diff` / `logdiff`) is
applied generically -- switching a transformation is a one-line change in
`00_config.R` and never touches this code.

```{r transform}
model_data <- transform_data(raw_data, config$variables)
knitr::kable(head(model_data, 6), digits = 4, caption = "Transformed data (first 6 months)")
```

# VAR estimation

Variables are split into endogenous/exogenous automatically from each
variable's `role` in config. Eleven monthly seasonal dummies (Feb-Dec; Jan
is the reference month) are always appended to the exogenous block to
control for calendar seasonality. The lag order follows
`config$lag_selection` (`fixed` or `auto` via `vars::VARselect()`).

```{r var-estimation}
var_out <- estimate_var(model_data, config)
cat(sprintf("Lag order p = %d (method: %s). Estimation sample: %d observations.\n",
            var_out$p, config$lag_selection$method, nrow(var_out$endo) - var_out$p))
```

# Diagnostics

Per the thesis's model-validation requirements: unit-root tests (ADF,
KPSS), a Johansen cointegration test, VAR residual diagnostics
(autocorrelation, normality), and companion-matrix stability. One compact
table per check -- read the statistics against the reported critical
values / p-values directly, no automatic interpretation.

```{r unit-root-level}
knitr::kable(unit_root_table(raw_data, config$variables, space = "level"), digits = 3,
             caption = "Unit-root tests, level (ADF H0: unit root; KPSS H0: stationary; p < 0.05 rejects H0)")
```

```{r unit-root-transform}
knitr::kable(unit_root_table(raw_data, config$variables, space = "transform"), digits = 3,
             caption = "Unit-root tests, after transform (ADF H0: unit root; KPSS H0: stationary; p < 0.05 rejects H0)")
```

## Deterministic components and unit roots

```{r stationarity-table}
stationarity <- stationarity_table(raw_data, config)
knitr::kable(stationarity, digits = 3, caption = stationarity_note(stationarity))
```

```{r za-table}
knitr::kable(za_table(raw_data, config), digits = 3,
             caption = "Zivot-Andrews, endogenous level shift (H0: unit root with no break)")
```

```{r johansen-test}
knitr::kable(johansen_test(raw_data, config$variables, var_out$p), digits = 3,
             caption = "Johansen trace test (H0: cointegration rank <= r), 5% critical value")
```

```{r residual-diagnostics}
knitr::kable(residual_diagnostics(var_out$fit, var_out$p), digits = 3,
             caption = "VAR residual diagnostics")
```

```{r stability-table}
knitr::kable(stability_table(var_out$fit), digits = 3,
             caption = "Companion-matrix root moduli (sorted descending)")
```

# Identification

Structural shocks are identified via sign restrictions read generically
from `config$shocks` -- adding a shock or a restriction requires no changes
to `05_sign_restrictions.R`. For each of `config$identification$n_draws`
Haar-uniform random rotations $Q$, the structural impact matrix
$B_0 = P Q$ is checked against every configured shock's sign restrictions;
admissible draws are kept, and the **Fry-Pagan median-target** draw (the
single admissible model closest to the pointwise median IRF) is reported,
rather than the pointwise median itself.

Three shocks are labelled out of the $K = 4$ available, so identification is
**partial** throughout: an aggregate demand shock, a gas-specific shock, and
an electricity supply shock specific to the Italian market. Their restriction
sets are constructed to be **pairwise disjoint**, so that no admissible
rotation can satisfy two of them in the same column of $B_0$. This matters
more than it may appear: when two labelled shocks share a region of the sign
space, the same economic disturbance is assigned to different columns in
different draws, and the pooled distribution of impulse responses mixes them
-- the labelling problem of Fry & Pagan (2011, sec. 4). Two margins do the
separating.

The first is quantity. A disturbance that moves the electricity price and
electricity consumption in the *same* direction is a shift of demand along an
upward-sloping supply curve; one that moves them in *opposite* directions is
a shift of supply along a downward-sloping demand curve. The sign of
`energy_consumption` therefore separates the aggregate demand shock from the
two contractionary shocks, and is the standard supply/demand identification
argument (Kilian & Murphy, 2012).

The second is the relative price. Both contractionary shocks raise PUN, so
the level of the electricity price cannot tell them apart; what does is the
electricity-price response *net of* the gas-price response,
$\theta_{\text{pun}} - \theta_{\text{gas}}$ at impact. Under a gas-specific
shock PUN rises *because* TTF rises, and pass-through in logs is less than
one-for-one -- PUN also embeds the CO$_2$ price, non-fuel variable costs, and
the hours in which gas is not the marginal technology -- so the contrast is
negative. An electricity supply shock (plant outages, hydrological shortfall,
low renewable output, constrained imports) originates downstream of the gas
market, and Italy is a price taker on TTF, so PUN rises while the gas price
stays roughly flat and the contrast is positive. This is a restriction on log
responses, an elasticity comparison, and not a spark spread measured in
EUR/MWh; it restricts neither price level. A zero restriction on the gas
response would state the price-taking argument more directly, but requires
the mixed sign-and-zero algorithm of Arias, Rubio-Ramírez & Waggoner (2018)
in place of the Haar-uniform rotations used here.

Industrial production is left **unrestricted** under the electricity supply
shock. The macro cost of energy disturbances is the quantity this model is
built to measure, and restricting the sign of `ipi` would assume a
substantial part of that answer; the scales do not match either, since PUN is
Italian while the industrial production index is EU-wide. It is retained
under the gas-specific shock, where both the disturbance and the aggregate
are European.

Finally, a **sign normalisation**: a positive realisation of the
gas-specific or the electricity supply shock is an *adverse* disturbance,
raising price and lowering quantity. Impulse responses and historical
decomposition contributions below are to be read in that direction.

```{r identification}
ident <- identify_sign_restrictions(var_out$fit, config)

cat(sprintf(
  "Acceptance rate: %.2f%% (%d / %d rotations admissible).\n",
  100 * ident$draws$acceptance_rate, ident$draws$n_accepted, ident$draws$n_draws
))
```

The acceptance rate is a diagnostic of how informative the restrictions
are: a low rate means the restrictions rule out most of the rotation
space (informative, but a non-refutability check worth keeping in mind);
a rate near 100% means the restrictions barely bind.

```{r restrictions-table}
restr_table <- purrr::map_dfr(config$shocks, function(s) {
  purrr::imap_dfr(s$restrictions, function(r, var_id) {
    tibble::tibble(shock = s$name, variable = var_id, sign = r$sign,
                    horizons = paste(range(r$horizons), collapse = ":"))
  })
})
knitr::kable(restr_table, caption = "Configured sign restrictions")
```

# Bootstrap

Frequentist confidence bands (Inoue & Kilian, 2013): the reduced-form VAR
residuals are resampled, the sample is rebuilt recursively, the VAR is
re-estimated, and the **entire identification search above is re-run**
inside every replication -- not just a coefficient resample.

```{r bootstrap}
boot_out <- run_bootstrap(var_out$fit, var_out$endo, var_out$exo, var_out$p, config)

cat(sprintf(
  "Bootstrap: %d replications, %d succeeded, %d found no admissible rotation.\n",
  boot_out$n_boot, boot_out$n_ok, boot_out$n_failed
))
cat(sprintf(
  "Mean acceptance rate across successful replications: %.2f%%.\n",
  100 * mean(boot_out$boot_acceptance_rates)
))
```

```{r boot-accept-hist, fig.width=7, fig.height=4.5}
theme_par(mar = c(4, 4, 3, 1))
h <- hist(100 * boot_out$boot_acceptance_rates, breaks = 15, plot = FALSE)
## Drawn twice so the grid sits behind the bars: plot.histogram() has no
## panel.first.
plot(h, col = NA, border = NA, main = "Bootstrap replications: acceptance rate distribution",
     xlab = "Acceptance rate (%)", ylab = "Replications")
grid_h()
plot(h, col = pal[["skyblue"]], border = "white", add = TRUE)
```

# Results

Median-target impulse responses with `r sprintf("%.0f%%", 100 * config$bootstrap$conf_level)`
bootstrap confidence bands. Every endogenous variable enters the VAR in logs, so
the responses are percent deviations from baseline to a one-standard-deviation
structural shock (log points x 100; exactly, 100(e^theta - 1)).

```{r irf-plot-fn}
plot_irf_grid <- function(mt_irf, bands, var_names, shock_names, var_labels) {
  n_shocks <- length(shock_names)
  n_vars <- length(var_names)
  h <- seq_len(dim(mt_irf)[3]) - 1

  ## Bottom margin 4, not 3: the last row carries the "Horizon (months)"
  ## label at mgp[1] = 2.4 lines, which a 3-line margin clips.
  old_par <- theme_par(mfrow = c(n_vars, n_shocks), mar = c(4, 4, 3, 1))
  on.exit(par(old_par))

  for (v in seq_len(n_vars)) {
    for (s in seq_len(n_shocks)) {
      ## All endogenous variables enter the VAR in logs, so the responses are
      ## log points; x100 reads them as percent deviations from baseline.
      point <- 100 * mt_irf[v, s, ]
      lo <- 100 * bands$lower[v, s, ]
      hi <- 100 * bands$upper[v, s, ]

      plot(h, point, type = "n", ylim = range(c(point, lo, hi), na.rm = TRUE),
           xlab = if (v == n_vars) "Horizon (months)" else "",
           ylab = paste0(var_labels[v], " (%)"),
           main = if (v == 1) gsub("_", " ", shock_names[s]) else "",
           panel.first = grid_h())
      polygon(c(h, rev(h)), c(lo, rev(hi)),
              col = grDevices::adjustcolor(pal[["blue"]], alpha.f = 0.18), border = NA)
      abline(h = 0, lty = 2, col = "grey55")
      lines(h, point, col = pal[["blue"]], lwd = 2)
    }
  }
}

## config labels carry the RAW series' units ("PUN (EUR/MWh)"), which are wrong
## on an IRF axis: the responses are percent deviations, not levels. Strip the
## trailing unit parenthesis -- plot_irf_grid() appends "(%)" instead. Only the
## IRF grid uses these; the raw-series and ACF panels read v$label directly.
var_labels <- sub("\\s*\\(.*\\)$", "", purrr::map_chr(config$variables, "label"))
names(var_labels) <- purrr::map_chr(config$variables, "id")
```

```{r irf-plot, fig.width=9, fig.height=10}
plot_irf_grid(ident$mt$mt_irf, boot_out$bands, ident$var_names, ident$shock_names,
              var_labels[ident$var_names])
```

## Variance decomposition

Share of each variable's h-step forecast error variance attributable to the
identified shocks, for the same Fry-Pagan median-target model as the impulse
responses above. Horizons are months after impact, so h = 0 is the impact
period. Identification is **partial** -- only the shocks in `config$shocks`
are pinned down -- so the denominator is the full reduced-form forecast error
variance implied by $\Sigma_u$, not the sum over the identified columns of
$B_0$. The identified shares consequently sum to less than one, and the
`unidentified` column is the remainder left to the shocks the sign
restrictions do not identify. It is reported as-is: rescaling the shares to
sum to one would claim an identification this design does not deliver.

Brackets are `r sprintf("%.0f%%", 100 * config$bootstrap$conf_level)` pointwise
percentile bands over the same pooled bootstrap draws as the IRF bands -- the
FEVD is computed inside each replication, so no second bootstrap pass is run.

```{r fevd-table}
fevd_mt <- fevd_median_target(var_out$fit, ident, config$identification$horizon)
fevd_names <- c(gsub("_", " ", ident$shock_names), "Unidentified")

fevd_tbl <- purrr::imap_dfr(config$fevd$horizons, function(h, k) {
  tbl <- tibble::tibble(horizon = h, variable = unname(var_labels[ident$var_names]))
  for (b in seq_along(fevd_names)) {
    tbl[[fevd_names[b]]] <- sprintf("%.2f [%.2f, %.2f]", fevd_mt[, b, h + 1],
                                     boot_out$fevd_bands$lower[, b, k],
                                     boot_out$fevd_bands$upper[, b, k])
  }
  tbl
})
knitr::kable(fevd_tbl, caption = paste0(
  "Forecast error variance decomposition, median-target model, with ",
  sprintf("%.0f%%", 100 * config$bootstrap$conf_level), " bootstrap bands"))
```

```{r fevd-plot, fig.width=9, fig.height=7}
#' Stacked-area FEVD, one panel per variable: the identified shocks' shares
#' stacked from the bottom and the unidentified remainder filling the gap up
#' to one, so how much of each variable the restrictions actually explain is
#' the visible grey band rather than something the reader has to subtract.
plot_fevd_grid <- function(fevd, band_names, var_labels) {
  h <- seq_len(dim(fevd)[3]) - 1
  n_vars <- dim(fevd)[1]
  cols <- c(unname(pal[c("blue", "orange", "bluishgreen", "reddishpurple")])[
    seq_len(length(band_names) - 1)], "grey85")

  old_par <- theme_par(mfrow = c(ceiling(n_vars / 2), 2), mar = c(3.5, 4, 3, 1),
                       oma = c(0, 0, 3, 0))
  on.exit(par(old_par))

  for (v in seq_len(n_vars)) {
    cum <- rbind(0, apply(fevd[v, , ], 2, cumsum))
    plot(h, cum[nrow(cum), ], type = "n", ylim = c(0, 1),
         xlab = if (v > n_vars - 2) "Horizon (months)" else "",
         ylab = "Share of forecast error variance",
         main = var_labels[v], panel.first = grid_h())
    for (b in seq_along(band_names)) {
      polygon(c(h, rev(h)), c(cum[b, ], rev(cum[b + 1, ])), col = cols[b], border = NA)
    }
  }
  top_legend(legend = band_names, fill = cols, ncol = 2)
}

plot_fevd_grid(fevd_mt, fevd_names, var_labels[ident$var_names])
```

## Historical decomposition

Where the FEVD asks how much of the *average* forecast error each shock
explains, the historical decomposition asks what each shock did in each
individual month of the estimation sample. It uses the same Fry-Pagan
median-target model as the impulse responses and the variance decomposition
above: the structural shocks are recovered by inverting that single $B_0$
through the reduced-form residuals ($\varepsilon_t = B_0^{-1} u_t$) and
convolved with the structural moving-average weights
$\Theta_h = \Phi_h B_0$. Nothing is re-estimated and no rotation is redrawn.

Identification is **partial** here in exactly the sense it is in the FEVD.
$B_0$ is a full invertible $K \times K$ matrix, so all $K$ structural shocks
are recoverable for this rotation, but only the shocks in `config$shocks`
carry an economic label; the remaining columns are summed into a single
`other shocks` band rather than rescaled away or dropped.

The **baseline** is everything the structural shocks do not move: the
intercept, the 11 monthly seasonal dummies, the exogenous block (the
`energy_crisis` dummy and `res_capacity`), and the effect of the $p$
presample values. It is therefore **not** a pure trend -- in particular the
energy-crisis dummy's effect sits in the baseline, so the gas shock is
credited only with what that dummy leaves unexplained, and the 2022 bars
should be read against a baseline that has already shifted. Contributions of
all $K$ shocks sum to the plotted deviation by construction, so the bars add
up to the black line in every month.

No bootstrap bands and no per-episode contribution tables are reported for
the historical decomposition: it is the point-estimate path of the
median-target model.

```{r historical-decomposition}
hd <- historical_decomposition(var_out, ident)
cat(sprintf("Historical decomposition over %d months, %s to %s.\n",
            nrow(hd$y_dev), min(hd$dates), max(hd$dates)))
```

```{r hd-plot, fig.width=9, fig.height=10}
#' Stacked contribution bars over time, one panel per endogenous variable,
#' with the actual deviation from baseline drawn on top in black. barplot()
#' stacks mixed-sign columns wrong (every segment is laid out from zero
#' upward, so negative contributions overwrite positive ones), so the
#' contributions are split into their positive and negative parts and drawn
#' as two stacked barplots sharing one ylim; the segments then meet at zero
#' and their net total is the black line by construction.
plot_hd_grid <- function(hd, var_labels) {
  n_shocks <- length(hd$shock_names)
  n_vars <- dim(hd$contrib)[2]
  band_names <- c(gsub("_", " ", hd$shock_names), "Other Shocks")
  ## Same shock -> colour mapping as the FEVD panels, grey for the shocks
  ## the restrictions do not identify.
  cols <- c(unname(pal[c("blue", "orange", "bluishgreen", "reddishpurple")])[
    seq_len(n_shocks)], "grey85")
  at <- seq(1, length(hd$dates), by = 12)

  old_par <- theme_par(mfrow = c(n_vars, 1), mar = c(2.5, 4.5, 2.5, 1),
                       oma = c(0, 0, 3, 0))
  on.exit(par(old_par))

  for (v in seq_len(n_vars)) {
    ## All endogenous variables enter the VAR in logs, so x100 reads the
    ## contributions as percent deviations, exactly as in the IRF grid.
    m <- 100 * t(cbind(hd$contrib[, v, seq_len(n_shocks)],
                       rowSums(hd$contrib[, v, -seq_len(n_shocks), drop = FALSE])))
    pos <- pmax(m, 0)
    neg <- pmin(m, 0)

    ## Drawn empty first so the grid sits behind the bars (barplot() has no
    ## panel.first) and to take the bar midpoints from its return value.
    bp <- barplot(pos, space = 0, col = NA, border = NA,
                  ylim = range(colSums(pos), colSums(neg)),
                  ylab = "Percent deviation from baseline", main = var_labels[v])
    grid_h()
    barplot(pos, space = 0, col = cols, border = NA, add = TRUE, axes = FALSE)
    barplot(neg, space = 0, col = cols, border = NA, add = TRUE, axes = FALSE)
    abline(h = 0, col = "grey55")
    lines(bp, 100 * hd$y_dev[, v], col = "black", lwd = 1.2)
    axis(1, at = bp[at], labels = format(hd$dates[at], "%Y"))
  }
  top_legend(legend = c(band_names, "actual deviation"),
             fill = c(cols, NA), col = "black",
             lty = c(rep(NA, length(cols)), 1), lwd = c(rep(NA, length(cols)), 1.2),
             ncol = 3)
}

plot_hd_grid(hd, var_labels[ident$var_names])
```

## Conditional forecast

The conditional forecasts hold the reduced form and the admissible rotations
**fixed** at the 2006--2025 estimates above: nothing is re-estimated and no
rotation is redrawn on the 2026 data. Every series is reloaded to
`cf$data_end` with the same transformations; RES installed capacity, which
has no 2026 values, is carried forward by its last monthly step. The
one-step-ahead forecast error of month $t$ uses the fixed coefficients and
the *realised* lags, $u_t = y_t - \hat y_{t|t-1}$, and the structural shocks
are $\varepsilon_t = B_0^{-1} u_t$, in standard deviations. A variable
missing in the last data month is set to its expectation given that month's
observed errors, $E[u_m \mid u_o] = \Sigma_{mo} \Sigma_{oo}^{-1} u_o$: a
**nowcast**, not data.

Every forward path is the zero-residual forecast from the last data month
plus the effect of an $H \times K$ matrix $W$ of future structural shocks,
$\sum_{s \le t} \Theta_{t-s} w_s$ -- the historical decomposition's
convolution, run forward. Bands are
`r sprintf("%.0f%%", 100 * config$bootstrap$conf_level)` pointwise
percentiles over the admissible rotations: they measure identification
uncertainty alone, with the reduced form at its point estimate, and no
bootstrap is run here.

```{r conditional-forecast}
cf_out <- conditional_forecast(var_out, ident, hd, config)
pct_cf <- sprintf("%.0f%%", 100 * config$bootstrap$conf_level)
n_cf <- length(cf_out$dates)

cat(sprintf("Ex post: %s to %s. Forecast from %s: %s to %s (%d months).\n",
            format(min(cf_out$a_dates), "%b %Y"), format(max(cf_out$a_dates), "%b %Y"),
            format(cf_out$dates[n_cf], "%b %Y"), format(min(cf_out$fdates), "%b %Y"),
            format(max(cf_out$fdates), "%b %Y"), length(cf_out$fdates)))
if (length(cf_out$nowcast) > 0) {
  cat(sprintf("Nowcast, not data (%s): %s.\n", format(cf_out$dates[n_cf], "%b %Y"),
              paste(sprintf("%s = %.1f", cf_out$nowcast, exp(cf_out$y[n_cf, cf_out$nowcast])),
                    collapse = ", ")))
}
cat(sprintf(paste("Checks, max abs error: one-step error vs VAR residual %.1e;",
                  "TTF vs target %.1e; reference path across rotations %.1e.\n"),
            cf_out$checks[[1]], cf_out$checks[[2]], cf_out$checks[[3]]))
```

### Ex post: the 2026 shocks

The one-step-ahead structural shocks of the months after the estimation
sample in which all four variables are observed. The attack month
(`cf$attack_month`) is set against March 2022 and against the distribution
of $|\varepsilon|$ over the estimation sample, both for the median-target
model. Switching `cf$attack_shocks` off from the attack month on removes
their contribution, $\sum_{s \ge \text{attack}} \Theta_{t-s} \varepsilon_s$,
from the data.

```{r cf-shock-table}
knitr::kable(cf_out$shock_table, digits = 0, caption = paste0(
  "One-step-ahead structural shocks (s.d.): median-target model, ", pct_cf,
  " band over admissible rotations in brackets. Percentile: share of estimation-sample ",
  "months with a smaller |shock|, median-target model."))
```

```{r cf-plot-fns}
## Same shock -> colour mapping as the FEVD and historical-decomposition
## panels, grey for the shock the restrictions do not label.
cf_cols <- stats::setNames(
  c(unname(pal[c("blue", "orange", "bluishgreen", "reddishpurple")])[seq_along(ident$shock_names)],
    "grey85"),
  cf_out$shock_names)

#' One path with its rotation band behind it.
band_line <- function(x, mid, lo, hi, col) {
  polygon(c(x, rev(x)), c(lo, rev(hi)),
          col = grDevices::adjustcolor(col, alpha.f = 0.18), border = NA)
  lines(x, mid, col = col, lwd = 2)
}

#' One panel per structural shock: the median-target one-step shock of each
#' ex-post month (bars, s.d.) with its rotation band (whiskers). The shaded
#' month is the attack month.
plot_cf_shocks <- function(cf, cols) {
  x <- seq_along(cf$a_dates)
  a <- match(cf$attack, cf$a_dates)
  old_par <- theme_par(mfrow = c(2, 2), mar = c(3, 4, 3, 1))
  on.exit(par(old_par))

  for (j in seq_along(cf$shock_names)) {
    plot(range(x) + c(-0.5, 0.5), range(0, cf$eps$lower[, j], cf$eps$upper[, j]),
         type = "n", xaxt = "n", xlab = "", ylab = "Shock (s.d.)",
         main = gsub("_", " ", cf$shock_names[j]),
         panel.first = {
           rect(a - 0.5, -1e6, a + 0.5, 1e6, col = "grey93", border = NA)
           grid_h()
         })
    rect(x - 0.35, 0, x + 0.35, cf$eps$mt[, j], col = cols[j], border = NA)
    arrows(x, cf$eps$lower[, j], x, cf$eps$upper[, j], angle = 90, code = 3,
           length = 0.03, col = "grey30")
    abline(h = 0, col = "grey55")
    axis(1, at = x, labels = format(cf$a_dates, "%b"))
  }
}

#' One forward panel: the reference path dashed and each of `readings` with
#' its rotation band, all from the origin month. PUN and TTF in EUR/MWh after
#' the observed months since Jan 2025; consumption and IPI in % (100 x log)
#' deviation from the reference path.
plot_cf_panel <- function(cf, v, readings, cols, main) {
  n <- length(cf$dates)
  xf <- c(cf$dates[n], cf$fdates)
  level <- v %in% c("pun", "gas_price")
  past <- level & cf$dates >= as.Date("2025-01-01")
  path <- function(q, r) {
    p <- q[, v, r]
    if (level) exp(c(cf$y[n, v], p)) else c(0, 100 * (p - cf$paths$mt[, v, "reference"]))
  }

  plot(range(cf$dates[past], xf),
       range(exp(cf$y[past, v]), path(cf$paths$mt, "reference"),
             sapply(readings, path, q = cf$paths$lower),
             sapply(readings, path, q = cf$paths$upper)),
       type = "n", xlab = "", ylab = if (level) "EUR/MWh" else "% deviation from reference",
       main = main, panel.first = grid_h())
  for (r in readings) {
    band_line(xf, path(cf$paths$mt, r), path(cf$paths$lower, r), path(cf$paths$upper, r),
              cols[[r]])
  }
  ## Explicit black: theme_par()'s `fg` also sets the default line colour.
  lines(xf, path(cf$paths$mt, "reference"), lty = 2, lwd = 1.5, col = "black")
  if (level) lines(cf$dates[past], exp(cf$y[past, v]), lwd = 1.5, col = "black")
}
```

```{r cf-shocks-plot, fig.width=9, fig.height=6}
plot_cf_shocks(cf_out, cf_cols)
```

```{r cf-pun-plot, fig.width=7, fig.height=4.5}
cf_col <- cf_cols[[config$cf$attack_shocks[1]]]
theme_par(mar = c(3, 4.5, 3, 1), oma = c(0, 0, 3, 0))
with(cf_out$pun_cf, {
  plot(cf_out$a_dates, actual, type = "n", ylim = range(actual, lower, upper),
       xlab = "", ylab = "EUR/MWh", main = "PUN: actual and without the attack shocks",
       panel.first = grid_h())
  abline(v = cf_out$attack, lty = 2, col = "grey55")
  band_line(cf_out$a_dates, mt, lower, upper, cf_col)
  lines(cf_out$a_dates, actual, lwd = 2, col = "black")
})
top_legend(legend = c("actual", "without the attack shocks"), col = c("black", cf_col),
           lwd = 2, ncol = 2)
```

```{r cf-gap-table}
knitr::kable(cf_out$gap_table, caption = paste0(
  "PUN (EUR/MWh) from the attack month on: actual, without the attack shocks, and their ",
  "contribution (the difference); median-target model, ", pct_cf,
  " band over admissible rotations in brackets."))
```

### Persistence (D1)

TTF is held at its last observed level to `cf$horizon_end`, and the gap $r_t$
between that target and the zero-residual forecast is closed three ways. With
all shocks free, the minimum-norm shocks $w = R'(RR')^{-1} r$ (Waggoner & Zha,
1999), where $R$ collects $\Theta_{t-s}[\text{TTF}, j]$ over months $s \le t$
and shocks $j$, give the **reference path**: the path is the same under every
admissible rotation, the implied shocks are not. With one shock alone --
gas-specific, or industrial demand -- the shocks solve a lower-triangular
system month by month, every other shock at zero. That system divides by the
impact response of TTF to the shock, which the sign restriction bounds only at
zero; in the admissible rotations where it is nearly zero the implied shocks
explode, so a single-shock reading enters its band only in the rotations where
it is numerically defined (TTF on the target and every path finite). PUN and
TTF are in EUR/MWh; consumption and IPI in percent (100 x log) deviation from
the reference path.

```{r cf-d1-plot, fig.width=9, fig.height=7}
cat(sprintf(paste("Single-shock readings numerically defined in %d (gas-specific) and %d",
                  "(industrial demand) of %d admissible rotations.\n"),
            cf_out$n_defined[["gas"]], cf_out$n_defined[["demand"]], cf_out$n_rotations))

d1_cols <- c(gas = cf_cols[["Gas-Specific_Shock"]], demand = cf_cols[["Industrial_Demand_Shock"]])
theme_par(mfrow = c(2, 2), mar = c(3, 4.5, 3, 1), oma = c(0, 0, 3, 0))
for (v in c("pun", "gas_price", "energy_consumption", "ipi")) {
  plot_cf_panel(cf_out, v, names(d1_cols), d1_cols, var_labels[[v]])
}
top_legend(legend = c("observed", "reference (all shocks)", "gas-specific only",
                      "industrial demand only"),
           col = c("black", "black", d1_cols), lty = c(1, 2, 1, 1), lwd = c(1.5, 1.5, 2, 2),
           ncol = 4)
```

### Escalation stress test (C1) and comparison

C1 replays the median-target gas-specific shocks of `cf$c1_window`, from the
month after the origin and with every other shock at zero, on top of the
reference path; in the band, each rotation replays its own shocks of that
window. The table compares the four forward paths.

```{r cf-c1-plot, fig.width=7, fig.height=4.5}
c1_col <- c(c1 = cf_cols[["Gas-Specific_Shock"]])
theme_par(mar = c(3, 4.5, 3, 1), oma = c(0, 0, 3, 0))
plot_cf_panel(cf_out, "pun", "c1", c1_col, "PUN: escalation stress test (C1)")
top_legend(legend = c("observed", "reference (all shocks)", "escalation (C1)"),
           col = c("black", "black", c1_col), lty = c(1, 2, 1), lwd = c(1.5, 1.5, 2), ncol = 3)
```

```{r cf-comparison}
knitr::kable(cf_out$comparison, caption = paste0(
  "Forward paths, ", format(min(cf_out$fdates), "%b %Y"), " to ",
  format(max(cf_out$fdates), "%b %Y"), ": median-target model, ", pct_cf,
  " band over admissible rotations in brackets (none where every rotation agrees). ",
  "PUN in EUR/MWh; IPI and consumption in % (100 x log) deviation from the reference path in ",
  format(max(cf_out$fdates), "%b %Y"), "; peak month of the median-target path."))
```

## Saved output

```{r save-output}
saveRDS(list(config = config, var_out = var_out, ident = ident, boot_out = boot_out,
             fevd_mt = fevd_mt, hd = hd, cf = cf_out),
        file.path(config$paths$output_dir, "svar_results.rds"))
cat("Saved results to", file.path(config$paths$output_dir, "svar_results.rds"), "\n")
```
