---
title: "SVAR Analysis of the Italian Electricity Market"
subtitle: "Identification by Sign Restrictions (Kilian & Lütkepohl, 2017, Ch. 13)"
author: "Alessandro Macchi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")

source("00_config.R")
source("01_load_data.R")
source("02_transform.R")
source("03_sanity_checks.R")
source("04_var_model.R")
source("05_sign_restrictions.R")
source("06_bootstrap.R")

dir.create(config$paths$output_dir, showWarnings = FALSE, recursive = TRUE)
```

# Overview

This report estimates a reduced-form VAR for the Italian electricity market
and identifies structural shocks via **sign restrictions**, following Kilian
& Lütkepohl (2017, Ch. 13): Haar-uniform random rotations of the Cholesky
factor (via QR decomposition), an admissible-set search over the configured
sign restrictions, and Fry & Pagan (2011) median-target (MT) selection for
the reported impulse responses. Confidence bands come from a **frequentist**
residual bootstrap (Inoue & Kilian, 2013) that re-runs the entire
identification search inside every replication.

Every modeling choice below (variables, transformations, lags, restrictions,
bootstrap settings) is read from `00_config.R` -- that is the only file that
needs editing to change the analysis.

## Endogenous variables

```{r vars-table}
var_table <- purrr::map_dfr(config$variables, ~ tibble::tibble(
  id = .x$id, file = .x$file, label = .x$label,
  transform = .x$transform, role = .x$role
))
knitr::kable(var_table, caption = "Variables configured in 00_config.R")
```

# Data loading

Each raw CSV is parsed with its own delimiter, date format, and decimal
mark (per-variable settings in `00_config.R`), then aligned onto one common
monthly grid so any month missing from a source file surfaces as an
explicit `NA` instead of silently shifting the other series out of
alignment.

```{r load-data}
raw_data <- load_all_series(config)
cat(sprintf("Loaded %d monthly observations, %s to %s.\n",
            nrow(raw_data), min(raw_data$date), max(raw_data$date)))
knitr::kable(head(raw_data, 6), digits = 2, caption = "Raw data (first 6 months)")
```

```{r plot-raw, fig.width=9, fig.height=6}
par(mfrow = c(2, 2), mar = c(3, 4, 2, 1))
for (v in config$variables) {
  plot(raw_data$date, raw_data[[v$id]], type = "l", col = "steelblue",
       xlab = "", ylab = v$label, main = v$label)
}
```

# Transformations

Each variable's `transform` field (`level` / `log` / `diff` / `logdiff`) is
applied generically -- switching a transformation is a one-line change in
`00_config.R` and never touches this code.

```{r transform}
model_data <- transform_data(raw_data, config$variables)
knitr::kable(head(model_data, 6), digits = 4, caption = "Transformed data (first 6 months)")
```

# VAR estimation

Variables are split into endogenous/exogenous automatically from each
variable's `role` in config. The lag order follows `config$lag_selection`
(`fixed` or `auto` via `vars::VARselect()`).

```{r var-estimation}
var_out <- estimate_var(model_data, config)
cat(sprintf("Lag order p = %d (method: %s). Estimation sample: %d observations.\n",
            var_out$p, config$lag_selection$method, nrow(var_out$endo) - var_out$p))
```

# Diagnostics

Per the thesis's model-validation requirements: unit-root tests (ADF,
KPSS), a Johansen cointegration test, VAR residual diagnostics
(autocorrelation, normality), and companion-matrix stability. One compact
table per check -- read the statistics against the reported critical
values / p-values directly, no automatic interpretation.

```{r unit-root-level}
knitr::kable(unit_root_table(raw_data, config$variables, space = "level"), digits = 3,
             caption = "Unit-root tests, level (ADF H0: unit root; KPSS H0: stationary; p < 0.05 rejects H0)")
```

```{r unit-root-transform}
knitr::kable(unit_root_table(raw_data, config$variables, space = "transform"), digits = 3,
             caption = "Unit-root tests, after transform (ADF H0: unit root; KPSS H0: stationary; p < 0.05 rejects H0)")
```

```{r johansen-test}
knitr::kable(johansen_test(raw_data, config$variables, var_out$p), digits = 3,
             caption = "Johansen trace test (H0: cointegration rank <= r), 5% critical value")
```

```{r residual-diagnostics}
knitr::kable(residual_diagnostics(var_out$fit, var_out$p), digits = 3,
             caption = "VAR residual diagnostics")
```

```{r stability-table}
knitr::kable(stability_table(var_out$fit), digits = 3,
             caption = "Companion-matrix root moduli (sorted descending)")
```

# Identification

Structural shocks are identified via sign restrictions read generically
from `config$shocks` -- adding a shock or a restriction requires no changes
to `05_sign_restrictions.R`. For each of `config$identification$n_draws`
Haar-uniform random rotations $Q$, the structural impact matrix
$B_0 = P Q$ is checked against every configured shock's sign restrictions;
admissible draws are kept, and the **Fry-Pagan median-target** draw (the
single admissible model closest to the pointwise median IRF) is reported,
rather than the pointwise median itself.

```{r identification}
ident <- identify_sign_restrictions(var_out$fit, config)

cat(sprintf(
  "Acceptance rate: %.2f%% (%d / %d rotations admissible).\n",
  100 * ident$draws$acceptance_rate, ident$draws$n_accepted, ident$draws$n_draws
))
```

The acceptance rate is a diagnostic of how informative the restrictions
are: a low rate means the restrictions rule out most of the rotation
space (informative, but a non-refutability check worth keeping in mind);
a rate near 100% means the restrictions barely bind.

```{r restrictions-table}
restr_table <- purrr::map_dfr(config$shocks, function(s) {
  purrr::imap_dfr(s$restrictions, function(r, var_id) {
    tibble::tibble(shock = s$name, variable = var_id, sign = r$sign,
                    horizons = paste(range(r$horizons), collapse = ":"))
  })
})
knitr::kable(restr_table, caption = "Configured sign restrictions")
```

# Bootstrap

Frequentist confidence bands (Inoue & Kilian, 2013): the reduced-form VAR
residuals are resampled, the sample is rebuilt recursively, the VAR is
re-estimated, and the **entire identification search above is re-run**
inside every replication -- not just a coefficient resample.

```{r bootstrap}
boot_out <- run_bootstrap(var_out$fit, var_out$endo, var_out$p, config)

cat(sprintf(
  "Bootstrap: %d replications, %d succeeded, %d found no admissible rotation.\n",
  boot_out$n_boot, boot_out$n_ok, boot_out$n_failed
))
cat(sprintf(
  "Mean acceptance rate across successful replications: %.2f%%.\n",
  100 * mean(boot_out$boot_acceptance_rates)
))
```

```{r boot-accept-hist, fig.width=6, fig.height=4}
hist(100 * boot_out$boot_acceptance_rates, breaks = 15, col = "steelblue",
     main = "Bootstrap replications: acceptance rate distribution",
     xlab = "Acceptance rate (%)")
```

# Results

Median-target impulse responses with `r sprintf("%.0f%%", 100 * config$bootstrap$conf_level)`
bootstrap confidence bands.

```{r irf-plot-fn}
plot_irf_grid <- function(mt_irf, bands, var_names, shock_names, var_labels) {
  n_shocks <- length(shock_names)
  n_vars <- length(var_names)
  h <- seq_len(dim(mt_irf)[3]) - 1

  old_par <- par(mfrow = c(n_vars, n_shocks), mar = c(3, 4, 2, 1))
  on.exit(par(old_par))

  for (v in seq_len(n_vars)) {
    for (s in seq_len(n_shocks)) {
      point <- mt_irf[v, s, ]
      lo <- bands$lower[v, s, ]
      hi <- bands$upper[v, s, ]

      plot(h, point, type = "n", ylim = range(c(point, lo, hi), na.rm = TRUE),
           xlab = "", ylab = var_labels[v],
           main = if (v == 1) shock_names[s] else "")
      polygon(c(h, rev(h)), c(lo, rev(hi)),
              col = grDevices::adjustcolor("steelblue", alpha.f = 0.25), border = NA)
      lines(h, point, col = "steelblue", lwd = 2)
      abline(h = 0, lty = 2, col = "grey40")
    }
  }
}

var_labels <- purrr::map_chr(config$variables, "label")
names(var_labels) <- purrr::map_chr(config$variables, "id")
```

```{r irf-plot, fig.width=9, fig.height=10}
plot_irf_grid(ident$mt$mt_irf, boot_out$bands, ident$var_names, ident$shock_names,
              var_labels[ident$var_names])
```

## Saved output

```{r save-output}
saveRDS(list(config = config, var_out = var_out, ident = ident, boot_out = boot_out),
        file.path(config$paths$output_dir, "svar_results.rds"))
cat("Saved results to", file.path(config$paths$output_dir, "svar_results.rds"), "\n")
```
