---
title: "First Visual Stationarity Checks"
author: "Thesis SVAR project"
date: "`r Sys.Date()`"
output: html_document
---

This notebook gives a first visual check of the variables that may enter the SVAR.

The goal is not to decide the final model yet. In this first step, we only compare:

- plots in levels;
- ACF and PACF plots, which help us see persistence.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5
)

plots_dir <- "plots"
dir.create(plots_dir, showWarnings = FALSE)

make_plot_name <- function(variable, plot_type) {
  variable_name <- tolower(variable)
  variable_name <- gsub("[^a-z0-9]+", "_", variable_name)
  variable_name <- gsub("^_|_$", "", variable_name)

  file.path(plots_dir, paste(variable_name, plot_type, sep = "_"))
}

save_plot <- function(file_name, plot_code) {
  png(file_name, width = 10, height = 5, units = "in", res = 300)
  on.exit(dev.off())
  force(plot_code)
}
```

## 1. Load The Data

```{r load-data}
data_dir <- "data"

parse_month <- function(x) {
  as.Date(paste0("01-", x), format = "%d-%m-%Y")
}

parse_euro_number <- function(x) {
  x <- gsub("\\.", "", x)
  x <- gsub(",", ".", x)
  as.numeric(x)
}

read_semicolon_series <- function(file, value_name, new_name) {
  raw <- read.csv(
    file.path(data_dir, file),
    sep = ";",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  data.frame(
    Date = parse_month(raw$Date),
    value = parse_euro_number(raw[[value_name]]),
    variable = new_name
  )
}

read_comma_series <- function(file, value_name, new_name) {
  raw <- read.csv(
    file.path(data_dir, file),
    sep = ",",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  data.frame(
    Date = parse_month(raw$Date),
    value = as.numeric(raw[[value_name]]),
    variable = new_name
  )
}

pun <- read_semicolon_series("pun.csv", "PUN (€/MWh)", "PUN")
gas <- read_semicolon_series("natural_gas.csv", "Natural gas Europe (€/MWh)", "Natural gas")
res <- read_comma_series("italy_res_capacity_monthly.csv", "RES Installed Capacity (MW)", "RES capacity")
hdd <- read_comma_series("hdd.csv", "heating_degree_days", "Heating degree days")
cdd <- read_comma_series("cdd.csv", "cooling_degree_days", "Cooling degree days")
cons <- read_semicolon_series("energy_consumption.csv", "Volumi MWh", "Electricity consumption")

all_data <- rbind(pun, gas, res, hdd, cdd, cons)

summary(all_data)
```

## 2. Keep The Common Monthly Sample

For a VAR or SVAR, all variables need to be observed over the same dates.
The common sample is from January 2005 to December 2025.

I also exclude 2026 here because the June 2026 electricity consumption observation looks incomplete.

```{r common-sample}
start_date <- as.Date("2005-01-01")
end_date <- as.Date("2025-12-01")

plot_data <- all_data[
  all_data$Date >= start_date &
    all_data$Date <= end_date,
]

plot_data$year <- as.numeric(format(plot_data$Date, "%Y"))
plot_data$month <- as.numeric(format(plot_data$Date, "%m"))
plot_data$month_name <- factor(
  months(plot_data$Date),
  levels = months(as.Date(paste0("2025-", sprintf("%02d", 1:12), "-01")))
)

table(plot_data$variable)
```

## 3. Variables In Levels

This section looks at the original variables.

For each variable, we plot:

- the time series;
- the ACF;
- the PACF.

If the time series has a clear trend, or if the ACF decays very slowly, the variable may not be stationary in levels.

```{r level-time-series}
variables <- unique(plot_data$variable)

for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  plot_level <- function() {
    plot(
      x$Date,
      x$value,
      type = "l",
      main = paste(v, "- level"),
      xlab = "Date",
      ylab = v,
      col = "steelblue",
      lwd = 2
    )
    grid()
  }

  plot_level()

  save_plot(
    make_plot_name(v, "level_time_series.png"),
    plot_level()
  )
}
```

```{r level-acf}
for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  plot_acf <- function() {
    acf(
      x$value,
      lag.max = 36,
      main = paste(v, "- ACF in levels")
    )
  }

  plot_acf()

  save_plot(
    make_plot_name(v, "acf_levels.png"),
    plot_acf()
  )
}
```

```{r level-pacf}
for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  plot_pacf <- function() {
    pacf(
      x$value,
      lag.max = 36,
      main = paste(v, "- PACF in levels")
    )
  }

  plot_pacf()

  save_plot(
    make_plot_name(v, "pacf_levels.png"),
    plot_pacf()
  )
}
```

## 4. Preliminary Transformations

This chunk prepares a first dataset for the SVAR analysis.

I log-transform PUN, natural gas, and electricity consumption.
I also add monthly dummies, using January as the reference month.

```{r preliminary-transformations}
svar_long <- plot_data

log_variables <- c("PUN", "Natural gas", "Electricity consumption")

if (any(svar_long$value[svar_long$variable %in% log_variables] <= 0, na.rm = TRUE)) {
  stop("Log transformation requires strictly positive values.")
}

svar_long$value[svar_long$variable %in% log_variables] <- log(
  svar_long$value[svar_long$variable %in% log_variables]
)

svar_long$variable[svar_long$variable == "PUN"] <- "log_pun"
svar_long$variable[svar_long$variable == "Natural gas"] <- "log_natural_gas"
svar_long$variable[svar_long$variable == "Electricity consumption"] <- "log_electricity_consumption"
svar_long$variable[svar_long$variable == "RES capacity"] <- "res_capacity"
svar_long$variable[svar_long$variable == "Heating degree days"] <- "heating_degree_days"
svar_long$variable[svar_long$variable == "Cooling degree days"] <- "cooling_degree_days"

svar_data <- reshape(
  svar_long[, c("Date", "variable", "value")],
  idvar = "Date",
  timevar = "variable",
  direction = "wide"
)

names(svar_data) <- sub("^value\\.", "", names(svar_data))
svar_data <- svar_data[order(svar_data$Date), ]
row.names(svar_data) <- NULL

svar_data$month <- factor(
  format(svar_data$Date, "%m"),
  levels = sprintf("%02d", 1:12),
  labels = month.abb
)

month_dummies <- model.matrix(~ month, data = svar_data)[, -1]
colnames(month_dummies) <- paste0("month_", month.abb[-1])

svar_data <- cbind(svar_data, month_dummies)
svar_data$month <- NULL
```

## 5. ADF Stationarity Tests

The Augmented Dickey-Fuller test has this null hypothesis:

- the variable has a unit root, so it is not stationary.

If the p-value is below 0.05, we reject the null and treat the variable as stationary.
This is still a preliminary check, not the final modelling decision.
I use 12 lags because the data are monthly.

```{r adf-tests}
if (!requireNamespace("tseries", quietly = TRUE)) {
  stop("Package 'tseries' is needed for the ADF test. Install it with install.packages('tseries').")
}

adf_variables <- names(svar_data)[
  names(svar_data) != "Date" &
    !grepl("^month_", names(svar_data))
]

adf_lag <- 12

adf_results <- do.call(
  rbind,
  lapply(adf_variables, function(v) {
    test <- tseries::adf.test(
      na.omit(svar_data[[v]]),
      k = adf_lag
    )

    data.frame(
      variable = v,
      lags = adf_lag,
      adf_statistic = as.numeric(test$statistic),
      p_value = test$p.value,
      conclusion = ifelse(
        test$p.value < 0.05,
        "Reject unit root: stationary",
        "Do not reject unit root: non-stationary"
      )
    )
  })
)

adf_results
```

## 6. First Differences For Non-Stationary Logged Variables

The previous ADF test suggests that `log_pun` and `log_electricity_consumption`
may still be non-stationary. As a next preliminary step, I take first differences
of these two logged variables and run the ADF test again.

```{r first-difference-adf-tests}
diff_variables <- c("log_pun", "log_electricity_consumption")

diff_data <- data.frame(
  Date = svar_data$Date[-1]
)

for (v in diff_variables) {
  diff_data[[paste0("d_", v)]] <- diff(svar_data[[v]])
}

diff_adf_results <- do.call(
  rbind,
  lapply(names(diff_data)[names(diff_data) != "Date"], function(v) {
    test <- tseries::adf.test(
      na.omit(diff_data[[v]]),
      k = adf_lag
    )

    data.frame(
      variable = v,
      lags = adf_lag,
      adf_statistic = as.numeric(test$statistic),
      p_value = test$p.value,
      conclusion = ifelse(
        test$p.value < 0.05,
        "Reject unit root: stationary",
        "Do not reject unit root: non-stationary"
      )
    )
  })
)

diff_adf_results
```
