---
title: "First Visual Stationarity Checks"
author: "Thesis SVAR project"
date: "`r Sys.Date()`"
output: html_document
---

This notebook gives a first visual check of the variables that may enter the SVAR.

The goal is not to decide the final model yet. In this first step, we only compare:

- plots in levels;
- ACF and PACF plots, which help us see persistence.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5
)
```

## 1. Load The Data

```{r load-data}
data_dir <- "data"

parse_month <- function(x) {
  as.Date(paste0("01-", x), format = "%d-%m-%Y")
}

parse_euro_number <- function(x) {
  x <- gsub("\\.", "", x)
  x <- gsub(",", ".", x)
  as.numeric(x)
}

read_semicolon_series <- function(file, value_name, new_name) {
  raw <- read.csv(
    file.path(data_dir, file),
    sep = ";",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  data.frame(
    Date = parse_month(raw$Date),
    value = parse_euro_number(raw[[value_name]]),
    variable = new_name
  )
}

read_comma_series <- function(file, value_name, new_name) {
  raw <- read.csv(
    file.path(data_dir, file),
    sep = ",",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  data.frame(
    Date = parse_month(raw$Date),
    value = as.numeric(raw[[value_name]]),
    variable = new_name
  )
}

pun <- read_semicolon_series("pun.csv", "PUN (€/MWh)", "PUN")
gas <- read_semicolon_series("natural_gas.csv", "Natural gas Europe (€/MWh)", "Natural gas")
res <- read_comma_series("italy_res_capacity_monthly.csv", "RES Installed Capacity (MW)", "RES capacity")
hdd <- read_comma_series("hdd.csv", "heating_degree_days", "Heating degree days")
cdd <- read_comma_series("cdd.csv", "cooling_degree_days", "Cooling degree days")
cons <- read_semicolon_series("energy_consumption.csv", "Volumi MWh", "Electricity consumption")

all_data <- rbind(pun, gas, res, hdd, cdd, cons)

summary(all_data)
```

## 2. Keep The Common Monthly Sample

For a VAR or SVAR, all variables need to be observed over the same dates.
The common sample is from January 2005 to December 2025.

I also exclude 2026 here because the June 2026 electricity consumption observation looks incomplete.

```{r common-sample}
start_date <- as.Date("2005-01-01")
end_date <- as.Date("2025-12-01")

plot_data <- all_data[
  all_data$Date >= start_date &
    all_data$Date <= end_date,
]

plot_data$year <- as.numeric(format(plot_data$Date, "%Y"))
plot_data$month <- as.numeric(format(plot_data$Date, "%m"))
plot_data$month_name <- factor(
  months(plot_data$Date),
  levels = months(as.Date(paste0("2025-", sprintf("%02d", 1:12), "-01")))
)

table(plot_data$variable)
```

## 3. Variables In Levels

This section looks at the original variables.

For each variable, we plot:

- the time series;
- the ACF;
- the PACF.

If the time series has a clear trend, or if the ACF decays very slowly, the variable may not be stationary in levels.

```{r level-time-series}
variables <- unique(plot_data$variable)

for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  plot(
    x$Date,
    x$value,
    type = "l",
    main = paste(v, "- level"),
    xlab = "Date",
    ylab = v,
    col = "steelblue",
    lwd = 2
  )
  grid()
}
```

```{r level-acf}
for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  acf(
    x$value,
    lag.max = 36,
    main = paste(v, "- ACF in levels")
  )
}
```

```{r level-pacf}
for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  pacf(
    x$value,
    lag.max = 36,
    main = paste(v, "- PACF in levels")
  )
}
```

## 5. Short Reading Guide

After looking at the plots, use this checklist:

- If the level plot has a clear trend, the variable is probably not stationary in levels.
- If the ACF decays slowly, the level variable may have a unit root.
- If the ACF has large spikes at lags 12, 24, or 36, the variable may have annual seasonality.
- If the PACF has a large first spike and the ACF decays slowly, the variable is highly persistent.

For the next step, we can add first differences or formal tests such as ADF and KPSS.
