---
title: "First Visual Stationarity Checks"
author: "Thesis SVAR project"
date: "`r Sys.Date()`"
output: html_document
---

This notebook gives a first visual check of the variables that may enter the SVAR.

The goal is not to decide the final model yet. In this first step, we only compare:

- plots in levels;
- ACF and PACF plots, which help us see persistence.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5
)

plots_dir <- "plots"
dir.create(plots_dir, showWarnings = FALSE)

make_plot_name <- function(variable, plot_type) {
  variable_name <- tolower(variable)
  variable_name <- gsub("[^a-z0-9]+", "_", variable_name)
  variable_name <- gsub("^_|_$", "", variable_name)

  file.path(plots_dir, paste(variable_name, plot_type, sep = "_"))
}

save_plot <- function(file_name, plot_code) {
  png(file_name, width = 10, height = 5, units = "in", res = 300)
  on.exit(dev.off())
  force(plot_code)
}
```

## 1. Load The Data

```{r load-data}
data_dir <- "data"

parse_month <- function(x) {
  as.Date(paste0("01-", x), format = "%d-%m-%Y")
}

parse_euro_number <- function(x) {
  x <- gsub("\\.", "", x)
  x <- gsub(",", ".", x)
  as.numeric(x)
}

read_semicolon_series <- function(file, value_name, new_name) {
  raw <- read.csv(
    file.path(data_dir, file),
    sep = ";",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  data.frame(
    Date = parse_month(raw$Date),
    value = parse_euro_number(raw[[value_name]]),
    variable = new_name
  )
}

read_comma_series <- function(file, value_name, new_name) {
  raw <- read.csv(
    file.path(data_dir, file),
    sep = ",",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  data.frame(
    Date = parse_month(raw$Date),
    value = as.numeric(raw[[value_name]]),
    variable = new_name
  )
}

pun <- read_semicolon_series("pun.csv", "PUN (€/MWh)", "PUN")
gas <- read_semicolon_series("natural_gas.csv", "Natural gas Europe (€/MWh)", "Natural gas")
res <- read_comma_series("italy_res_capacity_monthly.csv", "RES Installed Capacity (MW)", "RES capacity")
hdd <- read_comma_series("hdd.csv", "heating_degree_days", "Heating degree days")
cdd <- read_comma_series("cdd.csv", "cooling_degree_days", "Cooling degree days")
cons <- read_semicolon_series("energy_consumption.csv", "Volumi MWh", "Electricity consumption")

all_data <- rbind(pun, gas, res, hdd, cdd, cons)

summary(all_data)
```

## 2. Keep The Common Monthly Sample

For a VAR or SVAR, all variables need to be observed over the same dates.
The common sample is from January 2005 to December 2025.

I also exclude 2026 here because the June 2026 electricity consumption observation looks incomplete.

```{r common-sample}
start_date <- as.Date("2005-01-01")
end_date <- as.Date("2025-12-01")

plot_data <- all_data[
  all_data$Date >= start_date &
    all_data$Date <= end_date,
]

plot_data$year <- as.numeric(format(plot_data$Date, "%Y"))
plot_data$month <- as.numeric(format(plot_data$Date, "%m"))
plot_data$month_name <- factor(
  months(plot_data$Date),
  levels = months(as.Date(paste0("2025-", sprintf("%02d", 1:12), "-01")))
)

table(plot_data$variable)
```

## 3. Variables In Levels

This section looks at the original variables.

For each variable, we plot:

- the time series;
- the ACF;
- the PACF.

If the time series has a clear trend, or if the ACF decays very slowly, the variable may not be stationary in levels.

```{r level-time-series}
variables <- unique(plot_data$variable)

for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  plot_level <- function() {
    plot(
      x$Date,
      x$value,
      type = "l",
      main = paste(v, "- level"),
      xlab = "Date",
      ylab = v,
      col = "steelblue",
      lwd = 2
    )
    grid()
  }

  plot_level()

  save_plot(
    make_plot_name(v, "level_time_series.png"),
    plot_level()
  )
}
```

```{r level-acf}
for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  plot_acf <- function() {
    acf(
      x$value,
      lag.max = 36,
      main = paste(v, "- ACF in levels")
    )
  }

  plot_acf()

  save_plot(
    make_plot_name(v, "acf_levels.png"),
    plot_acf()
  )
}
```

```{r level-pacf}
for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  plot_pacf <- function() {
    pacf(
      x$value,
      lag.max = 36,
      main = paste(v, "- PACF in levels")
    )
  }

  plot_pacf()

  save_plot(
    make_plot_name(v, "pacf_levels.png"),
    plot_pacf()
  )
}
```

## 4. Preliminary Transformations

This chunk prepares a first dataset for the SVAR analysis.

I log-transform PUN, natural gas, and electricity consumption.
I also add monthly dummies, using January as the reference month.

```{r preliminary-transformations}
svar_long <- plot_data

log_variables <- c("PUN", "Natural gas", "Electricity consumption")

if (any(svar_long$value[svar_long$variable %in% log_variables] <= 0, na.rm = TRUE)) {
  stop("Log transformation requires strictly positive values.")
}

svar_long$value[svar_long$variable %in% log_variables] <- log(
  svar_long$value[svar_long$variable %in% log_variables]
)

svar_long$variable[svar_long$variable == "PUN"] <- "log_pun"
svar_long$variable[svar_long$variable == "Natural gas"] <- "log_natural_gas"
svar_long$variable[svar_long$variable == "Electricity consumption"] <- "log_electricity_consumption"
svar_long$variable[svar_long$variable == "RES capacity"] <- "res_capacity"
svar_long$variable[svar_long$variable == "Heating degree days"] <- "heating_degree_days"
svar_long$variable[svar_long$variable == "Cooling degree days"] <- "cooling_degree_days"

svar_data <- reshape(
  svar_long[, c("Date", "variable", "value")],
  idvar = "Date",
  timevar = "variable",
  direction = "wide"
)

names(svar_data) <- sub("^value\\.", "", names(svar_data))
svar_data <- svar_data[order(svar_data$Date), ]
row.names(svar_data) <- NULL

svar_data$month <- factor(
  format(svar_data$Date, "%m"),
  levels = sprintf("%02d", 1:12),
  labels = month.abb
)

month_dummies <- model.matrix(~ month, data = svar_data)[, -1]
colnames(month_dummies) <- paste0("month_", month.abb[-1])

svar_data <- cbind(svar_data, month_dummies)
svar_data$month <- NULL
```

## 5. ADF Stationarity Tests

The Augmented Dickey-Fuller test has this null hypothesis:

- the variable has a unit root, so it is not stationary.

If the p-value is below 0.05, we reject the null and treat the variable as stationary.
This is still a preliminary check, not the final modelling decision.
I use 12 lags because the data are monthly.

```{r adf-tests}
if (!requireNamespace("tseries", quietly = TRUE)) {
  stop("Package 'tseries' is needed for the ADF test. Install it with install.packages('tseries').")
}

adf_variables <- names(svar_data)[
  names(svar_data) != "Date" &
    !grepl("^month_", names(svar_data))
]

adf_lag <- 12

adf_results <- do.call(
  rbind,
  lapply(adf_variables, function(v) {
    test <- tseries::adf.test(
      na.omit(svar_data[[v]]),
      k = adf_lag
    )

    data.frame(
      variable = v,
      lags = adf_lag,
      adf_statistic = as.numeric(test$statistic),
      p_value = test$p.value,
      conclusion = ifelse(
        test$p.value < 0.05,
        "Reject unit root: stationary",
        "Do not reject unit root: non-stationary"
      )
    )
  })
)

adf_results
```

## 6. First Differences For Non-Stationary Logged Variables

The previous ADF test suggests that `log_pun` and `log_electricity_consumption`
may still be non-stationary. As a next preliminary step, I take first differences
of these two logged variables and run the ADF test again.

```{r first-difference-adf-tests}
diff_variables <- c("log_pun", "log_electricity_consumption")

diff_data <- data.frame(
  Date = svar_data$Date[-1]
)

for (v in diff_variables) {
  diff_data[[paste0("d_", v)]] <- diff(svar_data[[v]])
}

diff_adf_results <- do.call(
  rbind,
  lapply(names(diff_data)[names(diff_data) != "Date"], function(v) {
    test <- tseries::adf.test(
      na.omit(diff_data[[v]]),
      k = adf_lag
    )

    data.frame(
      variable = v,
      lags = adf_lag,
      adf_statistic = as.numeric(test$statistic),
      p_value = test$p.value,
      conclusion = ifelse(
        test$p.value < 0.05,
        "Reject unit root: stationary",
        "Do not reject unit root: non-stationary"
      )
    )
  })
)

diff_adf_results
```

## 7. Reduced-Form VAR

This section estimates a first standard VAR.

At this stage there are no structural restrictions yet. The model is only the
reduced-form VAR that will later be used as the starting point for the SVAR.

I use first differences for `log_pun` and `log_electricity_consumption`, because
the previous section suggested that these logged variables may be non-stationary.
The other variables are kept in the same form used above. This is still a
preliminary specification.

```{r reduced-form-var-data}
var_endogenous <- c(
  "d_log_pun",
  "log_natural_gas",
  "res_capacity",
  "heating_degree_days",
  "cooling_degree_days",
  "d_log_electricity_consumption"
)

var_model_data <- data.frame(
  Date = svar_data$Date[-1],
  d_log_pun = diff(svar_data$log_pun),
  log_natural_gas = svar_data$log_natural_gas[-1],
  res_capacity = svar_data$res_capacity[-1],
  heating_degree_days = svar_data$heating_degree_days[-1],
  cooling_degree_days = svar_data$cooling_degree_days[-1],
  d_log_electricity_consumption = diff(svar_data$log_electricity_consumption)
)

var_month_dummies <- svar_data[
  -1,
  grepl("^month_", names(svar_data)),
  drop = FALSE
]

var_model_data <- cbind(var_model_data, var_month_dummies)
var_model_data <- na.omit(var_model_data)

summary(var_model_data[, var_endogenous])
```

```{r reduced-form-var-functions}
make_var_matrices <- function(y, lags, exog = NULL, start_row = lags + 1) {
  rows <- start_row:nrow(y)

  y_dependent <- y[rows, , drop = FALSE]

  x_regressors <- matrix(1, nrow = length(rows), ncol = 1)
  colnames(x_regressors) <- "constant"

  for (lag in seq_len(lags)) {
    lagged_y <- y[rows - lag, , drop = FALSE]
    colnames(lagged_y) <- paste0(colnames(y), "_lag", lag)
    x_regressors <- cbind(x_regressors, lagged_y)
  }

  if (!is.null(exog)) {
    x_regressors <- cbind(x_regressors, exog[rows, , drop = FALSE])
  }

  list(y = y_dependent, x = x_regressors)
}

fit_var_ols <- function(y, lags, exog = NULL, start_row = lags + 1) {
  var_matrices <- make_var_matrices(
    y = y,
    lags = lags,
    exog = exog,
    start_row = start_row
  )

  beta <- qr.solve(var_matrices$x, var_matrices$y)
  residuals <- var_matrices$y - var_matrices$x %*% beta

  list(
    lags = lags,
    y = var_matrices$y,
    x = var_matrices$x,
    beta = beta,
    residuals = residuals,
    residual_covariance = crossprod(residuals) / nrow(residuals)
  )
}

compute_information_criteria <- function(var_fit) {
  n_obs <- nrow(var_fit$residuals)
  n_equations <- ncol(var_fit$y)
  n_parameters <- ncol(var_fit$x) * n_equations

  log_det_sigma <- as.numeric(
    determinant(var_fit$residual_covariance, logarithm = TRUE)$modulus
  )

  data.frame(
    lags = var_fit$lags,
    AIC = log_det_sigma + 2 * n_parameters / n_obs,
    BIC = log_det_sigma + log(n_obs) * n_parameters / n_obs,
    HQIC = log_det_sigma + 2 * log(log(n_obs)) * n_parameters / n_obs
  )
}
```

```{r reduced-form-var-lag-selection}
var_y <- as.matrix(var_model_data[, var_endogenous])

var_exog <- as.matrix(
  var_model_data[
    ,
    grepl("^month_", names(var_model_data)),
    drop = FALSE
  ]
)

max_lags <- 12
lag_selection_start <- max_lags + 1

lag_selection <- do.call(
  rbind,
  lapply(seq_len(max_lags), function(p) {
    var_fit_p <- fit_var_ols(
      y = var_y,
      lags = p,
      exog = var_exog,
      start_row = lag_selection_start
    )

    compute_information_criteria(var_fit_p)
  })
)

lag_selection
```

```{r reduced-form-var-estimation}
selected_lag <- lag_selection$lags[which.min(lag_selection$BIC)]

var_fit <- fit_var_ols(
  y = var_y,
  lags = 12, # CHANGE HERE FOR BIC SELECTION
  exog = var_exog
)

var_equation_fit <- data.frame(
  equation = colnames(var_fit$y),
  r_squared = 1 -
    colSums(var_fit$residuals^2) /
      colSums(scale(var_fit$y, scale = FALSE)^2),
  row.names = NULL
)

selected_lag
var_equation_fit
var_fit$residual_covariance
```

```{r reduced-form-var-stability}
var_stability <- function(var_fit) {
  n_variables <- ncol(var_fit$y)
  n_lags <- var_fit$lags

  companion <- matrix(0, n_variables * n_lags, n_variables * n_lags)

  for (lag in seq_len(n_lags)) {
    lag_rows <- (2 + (lag - 1) * n_variables):(1 + lag * n_variables)
    coefficient_matrix <- t(var_fit$beta[lag_rows, , drop = FALSE])

    companion[
      1:n_variables,
      ((lag - 1) * n_variables + 1):(lag * n_variables)
    ] <- coefficient_matrix
  }

  if (n_lags > 1) {
    companion[
      (n_variables + 1):(n_variables * n_lags),
      1:(n_variables * (n_lags - 1))
    ] <- diag(n_variables * (n_lags - 1))
  }

  roots <- eigen(companion)$values

  data.frame(
    largest_root_modulus = max(Mod(roots)),
    stable = max(Mod(roots)) < 1
  )
}

var_stability(var_fit)
```

## 8. Reduced-Form VAR Diagnostics

This section gives a compact residual check for the reduced-form VAR.

These diagnostics are not used to choose the structural identification. They are
only a quick warning system for residual autocorrelation, conditional
heteroskedasticity, and strong residual correlation across equations.

```{r compact-var-residual-diagnostics}
var_residuals <- as.data.frame(var_fit$residuals)
colnames(var_residuals) <- colnames(var_fit$y)

diagnostic_lag <- 12

arch_lm_test <- function(x, lags) {
  squared_residuals <- na.omit(x)^2
  rows <- (lags + 1):length(squared_residuals)

  y <- squared_residuals[rows]
  x_lags <- sapply(seq_len(lags), function(lag) {
    squared_residuals[rows - lag]
  })

  colnames(x_lags) <- paste0("squared_residual_lag", seq_len(lags))

  model <- lm(y ~ x_lags)
  statistic <- length(y) * summary(model)$r.squared
  p_value <- pchisq(statistic, df = lags, lower.tail = FALSE)

  data.frame(
    statistic = statistic,
    p_value = p_value
  )
}

residual_correlation <- cor(var_residuals)
max_abs_residual_correlation <- apply(
  abs(residual_correlation - diag(ncol(residual_correlation))),
  1,
  max
)

compact_diagnostics <- do.call(
  rbind,
  lapply(colnames(var_residuals), function(v) {
    ljung_box_test <- Box.test(
      var_residuals[[v]],
      lag = diagnostic_lag,
      type = "Ljung-Box"
    )

    arch_test <- arch_lm_test(
      var_residuals[[v]],
      lags = diagnostic_lag
    )

    data.frame(
      equation = v,
      residual_sd = sd(var_residuals[[v]]),
      ljung_box_p_value = ljung_box_test$p.value,
      arch_p_value = arch_test$p_value,
      max_abs_residual_correlation = max_abs_residual_correlation[v],
      row.names = NULL
    )
  })
)

compact_diagnostics
```

Small p-values in the Ljung-Box or ARCH columns indicate residual patterns that
the reduced-form VAR has not fully absorbed.

## 9. Recursive Identification Setup

This section prepares a possible Cholesky ordering for the future SVAR step.

No structural shocks are estimated here yet. The goal is only to write down the
recursive ordering clearly before implementing the SVAR.

The proposed ordering is:

1. `res_capacity`
2. `log_natural_gas`
3. `d_log_electricity_consumption`
4. `d_log_pun`

In words, this ordering means that variables earlier in the list may affect
variables later in the list contemporaneously, but variables later in the list do
not affect earlier variables within the same month.

```{r recursive-identification-ordering}
cholesky_order <- c(
  "res_capacity",
  "log_natural_gas",
  "log_electricity_consumption",
  "log_pun"
)

cholesky_order_check <- data.frame(
  position = seq_along(cholesky_order),
  variable = cholesky_order,
  available_in_var_data = cholesky_order %in% colnames(var_model_data)
)

cholesky_order_check
```

```{r recursive-identification-dataset}
if (!all(cholesky_order %in% colnames(var_model_data))) {
  stop("At least one variable in the Cholesky ordering is missing from var_model_data.")
}

cholesky_data <- var_model_data[, c("Date", cholesky_order)]

summary(cholesky_data[, cholesky_order])
```

The current reduced-form VAR in Step 7 also includes `heating_degree_days` and
`cooling_degree_days`. In the structural model below, I keep the recursive
ordering as a four-variable endogenous system and include the two weather
variables as exogenous regressors. This means that weather can help explain the
monthly dynamics of the endogenous variables, but no structural weather shock is
identified.

## 11. Structural VAR Estimation

This section estimates the first recursive Structural VAR using the `bsvars`
package.

The endogenous variables are the four variables in the Cholesky order. The
exogenous variables are:

- `heating_degree_days`;
- `cooling_degree_days`;
- monthly dummies.

The `bsvars` model uses the recursive lower-triangular identification for the
structural matrix `B`. The number of MCMC draws below is deliberately modest, so
that the notebook remains easy to run while we are still developing the thesis
specification.

```{r recursive-svar-bsvars-data}
if (!requireNamespace("bsvars", quietly = TRUE)) {
  stop("Package 'bsvars' is required for the SVAR estimation. Please install it first.")
}

svar_y <- as.matrix(cholesky_data[, cholesky_order])

svar_exogenous_variables <- c(
  "heating_degree_days",
  "cooling_degree_days",
  names(var_model_data)[grepl("^month_", names(var_model_data))]
)

svar_exog <- as.matrix(
  var_model_data[
    ,
    svar_exogenous_variables,
    drop = FALSE
  ]
)

svar_lags <- var_fit$lags

svar_reduced_form_fit <- fit_var_ols(
  y = svar_y,
  lags = svar_lags,
  exog = svar_exog
)

svar_dimensions <- data.frame(
  object = c("endogenous data", "exogenous data"),
  rows = c(nrow(svar_y), nrow(svar_exog)),
  columns = c(ncol(svar_y), ncol(svar_exog))
)

svar_exogenous_variables
svar_lags
svar_dimensions
```

```{r recursive-svar-bsvars-estimation}
set.seed(123)

svar_burn_in_draws <- 200
svar_posterior_draws <- 500
svar_thin <- 1

svar_specification <- bsvars::specify_bsvar$new(
  data = svar_y,
  p = svar_lags,
  exogenous = svar_exog
)

svar_burn_in <- bsvars::estimate(
  svar_specification,
  S = svar_burn_in_draws,
  thin = svar_thin,
  show_progress = FALSE
)

svar_posterior <- bsvars::estimate(
  svar_burn_in,
  S = svar_posterior_draws,
  thin = svar_thin,
  show_progress = FALSE
)

svar_B_normalisation <- diag(
  sign(diag(svar_posterior$last_draw$starting_values$B))
) %*% svar_posterior$last_draw$starting_values$B

if ("normalise_posterior" %in% getNamespaceExports("bsvars")) {
  bsvars::normalise_posterior(
    svar_posterior,
    svar_B_normalisation
  )
} else {
  svar_posterior <- bsvars::normalise(
    svar_posterior,
    svar_B_normalisation
  )
}

summary(svar_posterior)
```

```{r recursive-svar-impact-matrix}
svar_B_matrix <- svar_posterior$last_draw$starting_values$B
svar_impact_matrix <- solve(svar_B_matrix)

rownames(svar_impact_matrix) <- cholesky_order
colnames(svar_impact_matrix) <- paste0(cholesky_order, "_shock")

svar_impact_matrix
```

The impact matrix is lower triangular because of the recursive Cholesky
identification. This means that each variable can react within the same month to
its own shock and to shocks ordered before it. Because this is a Bayesian model,
the table reports the inverse of the last normalised draw of the structural
matrix `B`.

```{r recursive-svar-structural-shocks}
svar_structural_shock_draws <- bsvars::compute_structural_shocks(
  svar_posterior
)

svar_structural_shocks_median <- t(
  apply(
    svar_structural_shock_draws,
    c(1, 2),
    median
  )
)

svar_shock_dates <- cholesky_data$Date[
  (svar_lags + 1):nrow(cholesky_data)
]

svar_structural_shocks <- data.frame(
  Date = svar_shock_dates,
  svar_structural_shocks_median,
  row.names = NULL
)

names(svar_structural_shocks)[-1] <- paste0(cholesky_order, "_shock")

summary(svar_structural_shocks[, -1])
```

```{r recursive-svar-shock-covariance-check}
svar_structural_shock_covariance <- crossprod(
  as.matrix(svar_structural_shocks[, -1])
) / nrow(svar_structural_shocks)

svar_structural_shock_covariance
```

The covariance matrix of the structural shocks should be close to the identity
matrix. This is a check that the Cholesky transformation has converted the
correlated reduced-form residuals into orthogonal structural shocks.

## 16. Forecast Evaluation Against A Random Walk

This section compares the one-step-ahead forecast implied by the recursive SVAR
specification with a random walk benchmark.

For unconditional point forecasts, the recursive SVAR and its reduced-form VAR
produce the same point forecasts. The Cholesky identification changes how we
interpret shocks, but it does not change the baseline conditional mean forecast.
For this reason the expanding-window forecast below re-estimates the
reduced-form VAR equations with the same endogenous variables and exogenous
regressors used in the SVAR.

The target is PUN. Because the model uses `d_log_pun`, I evaluate one-step-ahead
forecasts for:

- `log_pun`, using RMSE and MAE;
- PUN in levels, using RMSE, MAE, and MAPE.

The random walk benchmark assumes that next period's `log_pun` remains equal to
the latest observed `log_pun`.

Because the model is estimated for `d_log_pun`, the back-transformation is:

`forecast(log_pun[t + 1]) = log_pun[t] + forecast(d_log_pun[t + 1])`.

The level forecast is then obtained as `exp(forecast(log_pun[t + 1]))`. This is
the median level forecast implied by the log forecast. A mean level forecast
would require an additional log-normal bias correction.

The SVAR forecast uses the realised next-period values of the exogenous
regressors, including `heating_degree_days` and `cooling_degree_days`. The
exercise should therefore be read as a conditional forecast comparison. For a
real-time forecasting exercise, those weather regressors would need to be
forecasted or lagged.

```{r forecast-evaluation-data}
forecast_data <- cholesky_data
forecast_data$log_pun <- svar_data$log_pun[
  match(forecast_data$Date, svar_data$Date)
]
forecast_data$pun <- exp(forecast_data$log_pun)

forecast_horizons <- 1
initial_window <- max(156, svar_lags + ncol(svar_reduced_form_fit$x) + 24)

forecast_design <- data.frame(
  validation_type = "expanding window",
  initial_window_months = initial_window,
  first_forecast_origin = forecast_data$Date[initial_window],
  last_available_date = forecast_data$Date[nrow(forecast_data)],
  forecast_horizon = forecast_horizons
)

forecast_design
```

```{r forecast-helper-functions}
forecast_var_path <- function(var_fit, y_history, exog_future, horizon) {
  forecasts <- matrix(
    NA,
    nrow = horizon,
    ncol = ncol(y_history)
  )

  colnames(forecasts) <- colnames(y_history)
  y_extended <- y_history

  for (step in seq_len(horizon)) {
    x_new <- c(constant = 1)

    for (lag in seq_len(var_fit$lags)) {
      lagged_values <- y_extended[nrow(y_extended) - lag + 1, ]
      names(lagged_values) <- paste0(names(lagged_values), "_lag", lag)
      x_new <- c(x_new, lagged_values)
    }

    x_new <- c(x_new, exog_future[step, ])
    x_new <- x_new[rownames(var_fit$beta)]

    forecasts[step, ] <- x_new %*% var_fit$beta
    y_extended <- rbind(y_extended, forecasts[step, ])
  }

  forecasts
}

compute_forecast_metrics <- function(actual, forecast) {
  error <- actual - forecast

  data.frame(
    RMSE = sqrt(mean(error^2, na.rm = TRUE)),
    MAE = mean(abs(error), na.rm = TRUE)
  )
}

compute_level_forecast_metrics <- function(actual, forecast) {
  error <- actual - forecast

  data.frame(
    RMSE = sqrt(mean(error^2, na.rm = TRUE)),
    MAE = mean(abs(error), na.rm = TRUE),
    MAPE = mean(abs(error / actual), na.rm = TRUE) * 100
  )
}
```

```{r expanding-window-forecast-evaluation}
forecast_results <- list()

for (horizon in forecast_horizons) {
  forecast_origins <- initial_window:(nrow(forecast_data) - horizon)

  horizon_results <- do.call(
    rbind,
    lapply(forecast_origins, function(origin) {
      train_y <- as.matrix(forecast_data[1:origin, cholesky_order])
      train_exog <- svar_exog[1:origin, , drop = FALSE]
      future_exog <- svar_exog[
        (origin + 1):(origin + horizon),
        ,
        drop = FALSE
      ]

      fit_at_origin <- fit_var_ols(
        y = train_y,
        lags = svar_lags,
        exog = train_exog
      )

      forecast_path <- forecast_var_path(
        var_fit = fit_at_origin,
        y_history = train_y,
        exog_future = future_exog,
        horizon = horizon
      )

      current_log_pun <- forecast_data$log_pun[origin]
      actual_log_pun <- forecast_data$log_pun[origin + horizon]

      svar_log_pun_forecast <- current_log_pun +
        sum(forecast_path[, "d_log_pun"])

      random_walk_log_pun_forecast <- current_log_pun

      actual_pun <- exp(actual_log_pun)
      svar_pun_forecast <- exp(svar_log_pun_forecast)
      random_walk_pun_forecast <- exp(random_walk_log_pun_forecast)

      data.frame(
        horizon = horizon,
        forecast_origin = forecast_data$Date[origin],
        target_date = forecast_data$Date[origin + horizon],
        actual_log_pun = actual_log_pun,
        svar_log_pun_forecast = svar_log_pun_forecast,
        random_walk_log_pun_forecast = random_walk_log_pun_forecast,
        actual_pun = actual_pun,
        svar_pun_forecast = svar_pun_forecast,
        random_walk_pun_forecast = random_walk_pun_forecast
      )
    })
  )

  forecast_results[[as.character(horizon)]] <- horizon_results
}

forecast_results <- do.call(rbind, forecast_results)
row.names(forecast_results) <- NULL

head(forecast_results)
```

```{r one-step-forecast-comparison-plot}
plot(
  forecast_results$target_date,
  forecast_results$actual_pun,
  type = "l",
  col = "black",
  lwd = 2,
  xlab = "Target date",
  ylab = "PUN",
  main = "One-step-ahead PUN forecast comparison"
)

lines(
  forecast_results$target_date,
  forecast_results$svar_pun_forecast,
  col = "steelblue",
  lwd = 2
)

lines(
  forecast_results$target_date,
  forecast_results$random_walk_pun_forecast,
  col = "firebrick",
  lwd = 2,
  lty = 2
)

legend(
  "topleft",
  legend = c("Actual PUN", "SVAR forecast", "Random walk forecast"),
  col = c("black", "steelblue", "firebrick"),
  lwd = c(2, 2, 2),
  lty = c(1, 1, 2),
  bty = "n"
)

grid()
```

```{r forecast-metric-tables}
log_forecast_metrics <- do.call(
  rbind,
  lapply(forecast_horizons, function(horizon) {
    horizon_results <- forecast_results[
      forecast_results$horizon == horizon,
    ]

    svar_metrics <- compute_forecast_metrics(
      actual = horizon_results$actual_log_pun,
      forecast = horizon_results$svar_log_pun_forecast
    )

    random_walk_metrics <- compute_forecast_metrics(
      actual = horizon_results$actual_log_pun,
      forecast = horizon_results$random_walk_log_pun_forecast
    )

    rbind(
      data.frame(horizon = horizon, model = "Recursive SVAR", svar_metrics),
      data.frame(horizon = horizon, model = "Random walk", random_walk_metrics)
    )
  })
)

level_forecast_metrics <- do.call(
  rbind,
  lapply(forecast_horizons, function(horizon) {
    horizon_results <- forecast_results[
      forecast_results$horizon == horizon,
    ]

    svar_metrics <- compute_level_forecast_metrics(
      actual = horizon_results$actual_pun,
      forecast = horizon_results$svar_pun_forecast
    )

    random_walk_metrics <- compute_level_forecast_metrics(
      actual = horizon_results$actual_pun,
      forecast = horizon_results$random_walk_pun_forecast
    )

    rbind(
      data.frame(horizon = horizon, model = "Recursive SVAR", svar_metrics),
      data.frame(horizon = horizon, model = "Random walk", random_walk_metrics)
    )
  })
)

log_forecast_metrics
level_forecast_metrics
```

```{r forecast-relative-performance}
relative_rmse <- merge(
  log_forecast_metrics[
    log_forecast_metrics$model == "Recursive SVAR",
    c("horizon", "RMSE", "MAE")
  ],
  log_forecast_metrics[
    log_forecast_metrics$model == "Random walk",
    c("horizon", "RMSE", "MAE")
  ],
  by = "horizon",
  suffixes = c("_svar", "_random_walk")
)

relative_rmse$svar_rmse_relative_to_random_walk <-
  relative_rmse$RMSE_svar / relative_rmse$RMSE_random_walk

relative_rmse$svar_mae_relative_to_random_walk <-
  relative_rmse$MAE_svar / relative_rmse$MAE_random_walk

relative_rmse
```

Values below 1 mean that the recursive SVAR forecast performs better than the
random walk benchmark for that metric and horizon.
