---
title: "Energy SVAR Analysis"
author: "Alessandro Macchi"
date: "`r Sys.Date()`"
output: html_document
---

This notebook is the readable report layer for the thesis SVAR project.

The reusable code lives in `R/`. The most important file for experimentation is
`R/02_variable_catalogue.R`: when a new candidate variable is added, its raw
file, raw column, active model variant, role, and current SVAR order should be
described there first.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5
)

project_scripts <- file.path(
  "R",
  c(
    "00_config.R",
    "01_io.R",
    "02_variable_catalogue.R",
    "03_prepare_data.R",
    "04_transformations.R",
    "05_diagnostics.R",
    "06_var_svar.R",
    "07_forecasting.R",
    "08_plotting.R"
  )
)

invisible(lapply(project_scripts, source))
ensure_project_dirs()
```

## 1. Variable Catalogue

This catalogue is the central place where candidate variables and the current
experiment are described.

To quickly try another specification, change:

- `model_variant`: `level`, `log`, `d_level`, or `d_log`;
- `role`: `endogenous`, `exogenous`, or `unused`;
- `svar_order`: Cholesky order, only for endogenous variables.

```{r variable-catalogue}
catalogue_view <- describe_variable_catalogue(variable_catalogue)
compact_table(catalogue_view)
```

## 2. Load The Data

The raw monthly series are loaded using the catalogue. This avoids one
hard-coded `read.csv()` call per variable in the notebook.

```{r load-data}
all_data <- load_catalogue_data(variable_catalogue)

data_coverage <- summarise_data_coverage(all_data)
compact_table(data_coverage)
```

## 3. Keep The Common Monthly Sample

For a VAR or SVAR, all variables need to be observed over the same dates.
The current common sample is from January 2005 to December 2025.

I also exclude 2026 here because the June 2026 electricity consumption
observation looks incomplete.

```{r common-sample}
prepared_data <- prepare_candidate_data(
  all_data,
  catalogue = variable_catalogue,
  config = experiment_config
)

plot_data <- prepared_data$plot_data
svar_data <- prepared_data$svar_data

sample_counts <- summarise_sample_counts(plot_data)
compact_table(sample_counts)
```

## 4. Variables In Levels

This section looks at the original variables.

For each variable, we plot:

- the time series;
- the ACF;
- the PACF.

If the time series has a clear trend, or if the ACF decays very slowly, the
variable may not be stationary in levels.

```{r level-diagnostics, fig.show='hold', out.width='32%', fig.width=5, fig.height=3.5}
plot_level_diagnostics(plot_data)
```

## 5. Preliminary Transformations

This step prepares a first dataset for the SVAR analysis.

The current experiment uses the `model_variant` selected in the catalogue.
Monthly dummies are also added, using January as the reference month.

```{r preliminary-transformations}
transformation_plan <- describe_variable_catalogue(variable_catalogue)

compact_table(transformation_plan)

preview_columns <- c("Date", get_active_model_columns(variable_catalogue))
compact_table(head(svar_data[, preview_columns]))
```

## 6. ADF Stationarity Tests

The Augmented Dickey-Fuller test has this null hypothesis:

- the variable has a unit root, so it is not stationary.

If the p-value is below 0.05, we reject the null and treat the variable as
stationary. This is still a preliminary check, not the final modelling decision.
I use 12 lags because the data are monthly.

```{r adf-tests}
adf_variables <- get_catalogue_model_columns(variable_catalogue)

adf_results <- run_adf_tests(
  svar_data,
  variables = adf_variables,
  lags = experiment_config$adf_lag
)

compact_table(adf_results)
```

## 7. First Differences For Active Differenced Variables

The previous ADF test suggests that some logged variables may still be
non-stationary. As a next preliminary step, I inspect the active catalogue
variables whose `model_variant` is `d_level` or `d_log`.

```{r first-difference-adf-tests}
diff_variables <- get_difference_columns(variable_catalogue)

diff_adf_results <- run_adf_tests(
  svar_data,
  variables = diff_variables,
  lags = experiment_config$adf_lag
)

compact_table(diff_adf_results)
```

## 8. Structural VAR With Cholesky Identification

This section estimates the current baseline SVAR.

The current Cholesky ordering is read from `svar_order` in the catalogue.

The VAR is estimated on January 2005 to December 2018.
The pseudo out-of-sample evaluation is January 2019 to December 2025.

The variables marked as exogenous in the catalogue, plus monthly dummies, enter
as external regressors. In the recursive forecast exercise, their realised
values are used for the next month, so the exercise is conditional on these
exogenous paths.

```{r svar-data-input}
svar_variables <- get_svar_variables(variable_catalogue)
exogenous_variables <- get_exogenous_variables(
  svar_data,
  catalogue = variable_catalogue,
  include_month_dummies = experiment_config$include_month_dummies
)

svar_input <- build_svar_input(
  svar_data,
  catalogue = variable_catalogue,
  config = experiment_config
)

model_samples <- make_model_samples(
  svar_input,
  svar_variables = svar_variables,
  exogenous_variables = exogenous_variables,
  config = experiment_config
)

estimation_sample <- model_samples$estimation_sample
test_sample <- model_samples$test_sample
endog_estimation <- model_samples$endog_estimation
exog_estimation <- model_samples$exog_estimation

model_variable_table <- make_model_variable_table(
  svar_variables,
  exogenous_variables
)

compact_table(model_variable_table)

sample_sizes <- summarise_sample_sizes(estimation_sample, test_sample)
compact_table(sample_sizes)
```

## 9. Lag Selection And Reduced-Form VAR

The information criteria are reported using `VARselect()` with `lag.max = 12`.
For this baseline, I keep 12 lags to allow annual seasonality in monthly data.

```{r lag-selection-and-var}
lag_selection <- select_var_lags(
  endog = endog_estimation,
  exog = exog_estimation,
  lag_max = experiment_config$lag_max
)

# compact_table(lag_selection$criteria_table)

selected_lag <- experiment_config$selected_lag
compact_table(data.frame(selected_lag = selected_lag))

var_fit <- fit_reduced_form_var(
  endog = endog_estimation,
  exog = exog_estimation,
  selected_lag = selected_lag
)

var_diagnostics <- summarise_var_diagnostics(var_fit, selected_lag)
compact_table(var_diagnostics)

# summary(var_fit)
```

## 10. Cholesky A-Matrix And SVAR Estimation

For an A-model in `vars::SVAR()`, the `NA` entries are estimated and the numeric
entries are restrictions.

For recursive Cholesky identification, exact identification requires
`K * (K - 1) / 2` zero restrictions above the diagonal. The helper below checks
this for the currently selected endogenous block.

```{r svar-cholesky}
amat <- make_cholesky_amat(svar_variables)
identification_check <- check_cholesky_identification(amat)

amat
compact_table(identification_check)

svar_fit <- fit_svar_cholesky(var_fit, amat)
svar_fit$A
```

## 11. Recursive One-Step PUN Forecast Evaluation

For point forecasts, the SVAR and the reduced-form VAR imply the same
conditional mean. The recursive exercise therefore re-estimates the reduced-form
VAR each month and then converts the one-step PUN forecasts back to levels.

The random walk benchmark is:

- next PUN level equals the last observed PUN level.

The Clark-West test compares the squared forecast errors of the SVAR and the
random walk with a nesting adjustment. The null hypothesis is equal predictive
accuracy.

```{r recursive-forecast-evaluation}
recursive_forecasts <- run_recursive_forecasts(
  data = svar_input,
  svar_variables = svar_variables,
  exogenous_variables = exogenous_variables,
  selected_lag = selected_lag,
  catalogue = variable_catalogue,
  config = experiment_config
)

pun_forecast_accuracy <- forecast_accuracy_table(
  recursive_forecasts,
  target_name = experiment_config$forecast_target_name
)

compact_table(pun_forecast_accuracy)
```
