---
title: "First Visual Stationarity Checks"
author: "Thesis SVAR project"
date: "`r Sys.Date()`"
output: html_document
---

This notebook gives a first visual check of the variables that may enter the SVAR.

The goal is not to decide the final model yet. In this first step, we only compare:

- plots in levels;
- ACF and PACF plots, which help us see persistence.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5
)

plots_dir <- "plots"
dir.create(plots_dir, showWarnings = FALSE)

make_plot_name <- function(variable, plot_type) {
  variable_name <- tolower(variable)
  variable_name <- gsub("[^a-z0-9]+", "_", variable_name)
  variable_name <- gsub("^_|_$", "", variable_name)

  file.path(plots_dir, paste(variable_name, plot_type, sep = "_"))
}

save_plot <- function(file_name, plot_code) {
  png(file_name, width = 10, height = 5, units = "in", res = 300)
  on.exit(dev.off())
  force(plot_code)
}
```

## 1. Load The Data

```{r load-data}
data_dir <- "data"

parse_month <- function(x) {
  as.Date(paste0("01-", x), format = "%d-%m-%Y")
}

parse_euro_number <- function(x) {
  x <- gsub("\\.", "", x)
  x <- gsub(",", ".", x)
  as.numeric(x)
}

read_semicolon_series <- function(file, value_name, new_name) {
  raw <- read.csv(
    file.path(data_dir, file),
    sep = ";",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  data.frame(
    Date = parse_month(raw$Date),
    value = parse_euro_number(raw[[value_name]]),
    variable = new_name
  )
}

read_comma_series <- function(file, value_name, new_name) {
  raw <- read.csv(
    file.path(data_dir, file),
    sep = ",",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  data.frame(
    Date = parse_month(raw$Date),
    value = as.numeric(raw[[value_name]]),
    variable = new_name
  )
}

pun <- read_semicolon_series("pun.csv", "PUN (€/MWh)", "PUN")
gas <- read_semicolon_series("natural_gas.csv", "Natural gas Europe (€/MWh)", "Natural gas")
res <- read_comma_series("italy_res_capacity_monthly.csv", "RES Installed Capacity (MW)", "RES capacity")
hdd <- read_comma_series("hdd.csv", "heating_degree_days", "Heating degree days")
cdd <- read_comma_series("cdd.csv", "cooling_degree_days", "Cooling degree days")
cons <- read_semicolon_series("energy_consumption.csv", "Volumi MWh", "Electricity consumption")

all_data <- rbind(pun, gas, res, hdd, cdd, cons)

summary(all_data)
```

## 2. Keep The Common Monthly Sample

For a VAR or SVAR, all variables need to be observed over the same dates.
The common sample is from January 2005 to December 2025.

I also exclude 2026 here because the June 2026 electricity consumption observation looks incomplete.

```{r common-sample}
start_date <- as.Date("2005-01-01")
end_date <- as.Date("2025-12-01")

plot_data <- all_data[
  all_data$Date >= start_date &
    all_data$Date <= end_date,
]

plot_data$year <- as.numeric(format(plot_data$Date, "%Y"))
plot_data$month <- as.numeric(format(plot_data$Date, "%m"))
plot_data$month_name <- factor(
  months(plot_data$Date),
  levels = months(as.Date(paste0("2025-", sprintf("%02d", 1:12), "-01")))
)

table(plot_data$variable)
```

## 3. Variables In Levels

This section looks at the original variables.

For each variable, we plot:

- the time series;
- the ACF;
- the PACF.

If the time series has a clear trend, or if the ACF decays very slowly, the variable may not be stationary in levels.

```{r level-time-series}
variables <- unique(plot_data$variable)

for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  plot_level <- function() {
    plot(
      x$Date,
      x$value,
      type = "l",
      main = paste(v, "- level"),
      xlab = "Date",
      ylab = v,
      col = "steelblue",
      lwd = 2
    )
    grid()
  }

  plot_level()

  save_plot(
    make_plot_name(v, "level_time_series.png"),
    plot_level()
  )
}
```

```{r level-acf}
for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  plot_acf <- function() {
    acf(
      x$value,
      lag.max = 36,
      main = paste(v, "- ACF in levels")
    )
  }

  plot_acf()

  save_plot(
    make_plot_name(v, "acf_levels.png"),
    plot_acf()
  )
}
```

```{r level-pacf}
for (v in variables) {
  x <- plot_data[plot_data$variable == v, ]

  plot_pacf <- function() {
    pacf(
      x$value,
      lag.max = 36,
      main = paste(v, "- PACF in levels")
    )
  }

  plot_pacf()

  save_plot(
    make_plot_name(v, "pacf_levels.png"),
    plot_pacf()
  )
}
```

## 4. Preliminary Transformations

This chunk prepares a first dataset for the SVAR analysis.

I log-transform PUN, natural gas, and electricity consumption.
I also add monthly dummies, using January as the reference month.

```{r preliminary-transformations}
svar_long <- plot_data

log_variables <- c("PUN", "Natural gas", "Electricity consumption")

if (any(svar_long$value[svar_long$variable %in% log_variables] <= 0, na.rm = TRUE)) {
  stop("Log transformation requires strictly positive values.")
}

svar_long$value[svar_long$variable %in% log_variables] <- log(
  svar_long$value[svar_long$variable %in% log_variables]
)

svar_long$variable[svar_long$variable == "PUN"] <- "log_pun"
svar_long$variable[svar_long$variable == "Natural gas"] <- "log_natural_gas"
svar_long$variable[svar_long$variable == "Electricity consumption"] <- "log_electricity_consumption"
svar_long$variable[svar_long$variable == "RES capacity"] <- "res_capacity"
svar_long$variable[svar_long$variable == "Heating degree days"] <- "heating_degree_days"
svar_long$variable[svar_long$variable == "Cooling degree days"] <- "cooling_degree_days"

svar_data <- reshape(
  svar_long[, c("Date", "variable", "value")],
  idvar = "Date",
  timevar = "variable",
  direction = "wide"
)

names(svar_data) <- sub("^value\\.", "", names(svar_data))
svar_data <- svar_data[order(svar_data$Date), ]
row.names(svar_data) <- NULL

svar_data$month <- factor(
  format(svar_data$Date, "%m"),
  levels = sprintf("%02d", 1:12),
  labels = month.abb
)

month_dummies <- model.matrix(~ month, data = svar_data)[, -1]
colnames(month_dummies) <- paste0("month_", month.abb[-1])

svar_data <- cbind(svar_data, month_dummies)
svar_data$month <- NULL
```

## 5. ADF Stationarity Tests

The Augmented Dickey-Fuller test has this null hypothesis:

- the variable has a unit root, so it is not stationary.

If the p-value is below 0.05, we reject the null and treat the variable as stationary.
This is still a preliminary check, not the final modelling decision.
I use 12 lags because the data are monthly.

```{r adf-tests}
if (!requireNamespace("tseries", quietly = TRUE)) {
  stop("Package 'tseries' is needed for the ADF test. Install it with install.packages('tseries').")
}

adf_variables <- names(svar_data)[
  names(svar_data) != "Date" &
    !grepl("^month_", names(svar_data))
]

adf_lag <- 12

adf_results <- do.call(
  rbind,
  lapply(adf_variables, function(v) {
    test <- tseries::adf.test(
      na.omit(svar_data[[v]]),
      k = adf_lag
    )

    data.frame(
      variable = v,
      lags = adf_lag,
      adf_statistic = as.numeric(test$statistic),
      p_value = test$p.value,
      conclusion = ifelse(
        test$p.value < 0.05,
        "Reject unit root: stationary",
        "Do not reject unit root: non-stationary"
      )
    )
  })
)

adf_results
```

## 6. First Differences For Non-Stationary Logged Variables

The previous ADF test suggests that `log_pun` and `log_electricity_consumption`
may still be non-stationary. As a next preliminary step, I take first differences
of these two logged variables and run the ADF test again.

```{r first-difference-adf-tests}
diff_variables <- c("log_pun", "log_electricity_consumption")

diff_data <- data.frame(
  Date = svar_data$Date[-1]
)

for (v in diff_variables) {
  diff_data[[paste0("d_", v)]] <- diff(svar_data[[v]])
}

diff_adf_results <- do.call(
  rbind,
  lapply(names(diff_data)[names(diff_data) != "Date"], function(v) {
    test <- tseries::adf.test(
      na.omit(diff_data[[v]]),
      k = adf_lag
    )

    data.frame(
      variable = v,
      lags = adf_lag,
      adf_statistic = as.numeric(test$statistic),
      p_value = test$p.value,
      conclusion = ifelse(
        test$p.value < 0.05,
        "Reject unit root: stationary",
        "Do not reject unit root: non-stationary"
      )
    )
  })
)

diff_adf_results
```

## 7. Structural VAR With Cholesky Identification

This section estimates the thesis baseline SVAR.

The Cholesky ordering is:

1. `log_natural_gas`;
2. `d_log_energy_consumption`;
3. `d_log_pun`.

The VAR is estimated on January 2010 to December 2018.
The pseudo out-of-sample evaluation is January 2019 to December 2024.

The weather variables, renewable capacity, and monthly dummies enter as
exogenous variables. In the recursive forecast exercise, their realised values
are used for the next month, so the exercise is conditional on these exogenous
paths.

```{r svar-data-input}
if (!requireNamespace("vars", quietly = TRUE)) {
  stop("Package 'vars' is needed for the VAR and SVAR models. Install it with install.packages('vars').")
}

svar_variables <- c(
  "log_natural_gas",
  "d_log_energy_consumption",
  "d_log_pun"
)

monthly_dummy_variables <- grep("^month_", names(svar_data), value = TRUE)

exogenous_variables <- c(
  "heating_degree_days",
  "cooling_degree_days",
  "res_capacity",
  monthly_dummy_variables
)

svar_model_data <- svar_data[order(svar_data$Date), ]

# Data input placeholder for the final cleaned modelling dataset.
# If the data cleaning step is later moved to a separate script, replace
# this object with the imported final dataset with the same column names.
svar_model_data$d_log_energy_consumption <- c(
  NA,
  diff(svar_model_data$log_electricity_consumption)
)
svar_model_data$d_log_pun <- c(
  NA,
  diff(svar_model_data$log_pun)
)

required_svar_columns <- c(
  "Date",
  "log_natural_gas",
  "log_electricity_consumption",
  "log_pun",
  svar_variables,
  exogenous_variables
)

missing_svar_columns <- setdiff(required_svar_columns, names(svar_model_data))

if (length(missing_svar_columns) > 0) {
  stop(
    "The SVAR dataset is missing these columns: ",
    paste(missing_svar_columns, collapse = ", ")
  )
}

svar_input <- svar_model_data[
  complete.cases(svar_model_data[, required_svar_columns]),
  required_svar_columns
]

estimation_start <- as.Date("2005-01-01")
estimation_end <- as.Date("2018-12-01")
test_start <- as.Date("2019-01-01")
test_end <- as.Date("2025-12-01")

estimation_sample <- svar_input[
  svar_input$Date >= estimation_start & svar_input$Date <= estimation_end,
]

test_sample <- svar_input[
  svar_input$Date >= test_start & svar_input$Date <= test_end,
]

if (nrow(estimation_sample) == 0) {
  stop("The estimation sample is empty.")
}

if (nrow(test_sample) == 0) {
  stop("The test sample is empty.")
}

endog_estimation <- as.matrix(estimation_sample[, svar_variables])
exog_estimation <- as.matrix(estimation_sample[, exogenous_variables])

sample_sizes <- data.frame(
  sample = c("estimation", "test"),
  start = c(min(estimation_sample$Date), min(test_sample$Date)),
  end = c(max(estimation_sample$Date), max(test_sample$Date)),
  observations = c(nrow(estimation_sample), nrow(test_sample))
)

sample_sizes
```

## 8. Lag Selection And Reduced-Form VAR

The lag order is selected by Schwarz criterion, using `VARselect()` with
`lag.max = 12`.

```{r lag-selection-and-var}
lag_selection <- vars::VARselect(
  y = endog_estimation,
  lag.max = 12,
  type = "const",
  exogen = exog_estimation
)

lag_selection$criteria

selected_lag <- 12 # as.integer(lag_selection$selection[["AIC(n)"]])
selected_lag # 12 for annual seasonality

var_fit <- vars::VAR(
  y = endog_estimation,
  p = selected_lag,
  type = "const",
  exogen = exog_estimation
)

diagnostic_tests <- list(
  Portmanteau = vars::serial.test(
    var_fit,
    lags.pt = selected_lag + 12,
    type = "PT.asymptotic"
  )$serial,
  "Jarque-Bera" = vars::normality.test(
    var_fit,
    multivariate.only = TRUE
  )$jb.mul$JB,
  ARCH = vars::arch.test(
    var_fit,
    lags.multi = 5,
    multivariate.only = TRUE
  )$arch.mul
)

diagnostic_p_values <- sapply(
  diagnostic_tests,
  function(test) as.numeric(test$p.value)
)

var_diagnostics <- data.frame(
  test = names(diagnostic_tests),
  statistic = round(
    sapply(diagnostic_tests, function(test) as.numeric(test$statistic)),
    3
  ),
  df = sapply(
    diagnostic_tests,
    function(test) as.numeric(test$parameter)
  ),
  p_value = signif(diagnostic_p_values, 3),
  decision_5pct = ifelse(
    diagnostic_p_values < 0.05,
    "reject H0",
    "do not reject H0"
  )
)

var_diagnostics

# summary(var_fit)
```

## 9. Cholesky A-Matrix And SVAR Estimation

For an A-model in `vars::SVAR()`, the `NA` entries are estimated and the
numeric entries are restrictions.

With three variables, exact identification requires
`K * (K - 1) / 2 = 3` restrictions. These are the three zero restrictions above
the diagonal. The three lower off-diagonal entries are the contemporaneous
Cholesky transmission parameters.

```{r svar-cholesky}
k_endog <- length(svar_variables)

amat <- matrix(NA, nrow = k_endog, ncol = k_endog)
rownames(amat) <- svar_variables
colnames(amat) <- svar_variables
amat[upper.tri(amat)] <- 0

needed_restrictions <- k_endog * (k_endog - 1) / 2
zero_restrictions <- sum(amat[upper.tri(amat)] == 0)
free_lower_off_diagonal <- sum(is.na(amat[lower.tri(amat)]))

identification_check <- data.frame(
  k_endog = k_endog,
  needed_zero_restrictions = needed_restrictions,
  imposed_zero_restrictions = zero_restrictions,
  free_lower_off_diagonal_parameters = free_lower_off_diagonal,
  exactly_identified = zero_restrictions == needed_restrictions
)

if (!identification_check$exactly_identified) {
  stop("The Cholesky A-matrix is not exactly identified.")
}

amat
identification_check

svar_fit <- vars::SVAR(
  x = var_fit,
  estmethod = "scoring",
  Amat = amat,
  Bmat = NULL
)

summary(svar_fit)
```

## 10. Recursive One-Step PUN Forecast Evaluation

For point forecasts, the SVAR and the reduced-form VAR imply the same conditional
mean. The recursive exercise therefore re-estimates the reduced-form VAR each
month and then converts the one-step PUN forecasts back to levels.

The random walk benchmark is:

- next PUN level equals the last observed PUN level.

The Diebold-Mariano test compares the squared forecast errors of the SVAR and
the random walk. The null hypothesis is equal predictive accuracy.

```{r recursive-forecast-evaluation}
forecast_indices <- which(
  svar_input$Date >= test_start & svar_input$Date <= test_end
)

forecast_one_step <- function(forecast_index) {
  forecast_date <- svar_input$Date[forecast_index]

  recursive_train <- svar_input[
    svar_input$Date >= estimation_start &
      svar_input$Date < forecast_date,
  ]

  recursive_endog <- as.matrix(recursive_train[, svar_variables])
  recursive_exog <- as.matrix(recursive_train[, exogenous_variables])

  recursive_var <- vars::VAR(
    y = recursive_endog,
    p = selected_lag,
    type = "const",
    exogen = recursive_exog
  )

  next_exog <- as.matrix(
    svar_input[forecast_index, exogenous_variables, drop = FALSE]
  )
  colnames(next_exog) <- exogenous_variables

  new_regressors <- data.frame(matrix(nrow = 1, ncol = 0))

  for (lag in seq_len(selected_lag)) {
    lagged_values <- svar_input[
      forecast_index - lag,
      svar_variables,
      drop = FALSE
    ]
    names(lagged_values) <- paste0(names(lagged_values), ".l", lag)
    new_regressors <- cbind(new_regressors, lagged_values)
  }

  new_regressors$const <- 1
  new_regressors <- cbind(new_regressors, as.data.frame(next_exog))

  endog_forecast <- sapply(
    svar_variables,
    function(variable) {
      coefficients <- coef(recursive_var$varresult[[variable]])
      coefficients[is.na(coefficients)] <- 0

      missing_regressors <- setdiff(names(coefficients), names(new_regressors))

      if (length(missing_regressors) > 0) {
        stop(
          "Missing recursive forecast regressors: ",
          paste(missing_regressors, collapse = ", ")
        )
      }

      sum(
        as.numeric(new_regressors[1, names(coefficients)]) *
          coefficients
      )
    }
  )

  previous_index <- forecast_index - 1

  data.frame(
    Date = forecast_date,
    forecast_pun = exp(
      svar_input$log_pun[previous_index] +
        endog_forecast[["d_log_pun"]]
    ),
    actual_pun = exp(svar_input$log_pun[forecast_index]),
    rw_pun = exp(svar_input$log_pun[previous_index])
  )
}

recursive_forecasts <- do.call(
  rbind,
  lapply(forecast_indices, forecast_one_step)
)

forecast_metrics <- function(actual, model_forecast, rw_forecast) {
  model_error <- actual - model_forecast
  rw_error <- actual - rw_forecast

  rmse_model <- sqrt(mean(model_error^2, na.rm = TRUE))
  rmse_rw <- sqrt(mean(rw_error^2, na.rm = TRUE))

  data.frame(
    rmse_svar = rmse_model,
    rmse_rw = rmse_rw,
    mae_svar = mean(abs(model_error), na.rm = TRUE),
    mae_rw = mean(abs(rw_error), na.rm = TRUE),
    theils_u = rmse_model / rmse_rw
  )
}

clark_west_test <- function(actual, model_forecast, rw_forecast) {
  model_error <- actual - model_forecast
  rw_error <- actual - rw_forecast
  adjustment <- (model_forecast - rw_forecast)^2
  cw_series <- rw_error^2 - (model_error^2 - adjustment)
  cw_series <- cw_series[is.finite(cw_series)]

  if (length(cw_series) < 2 || sd(cw_series) == 0) {
    return(
      data.frame(
        clark_west_statistic = NA_real_,
        clark_west_p_value_one_sided = NA_real_
      )
    )
  }

  statistic <- mean(cw_series) / (sd(cw_series) / sqrt(length(cw_series)))

  data.frame(
    clark_west_statistic = statistic,
    clark_west_p_value_one_sided = 1 - pnorm(statistic)
  )
}

diebold_mariano_test <- function(actual, model_forecast, rw_forecast) {
  model_error <- actual - model_forecast
  rw_error <- actual - rw_forecast
  loss_difference <- model_error^2 - rw_error^2
  loss_difference <- loss_difference[is.finite(loss_difference)]

  if (length(loss_difference) < 2 || sd(loss_difference) == 0) {
    return(
      data.frame(
        dm_statistic = NA_real_,
        dm_p_value_two_sided = NA_real_,
        dm_p_value_svar_lower_loss = NA_real_
      )
    )
  }

  statistic <- mean(loss_difference) /
    (sd(loss_difference) / sqrt(length(loss_difference)))

  data.frame(
    dm_statistic = statistic,
    dm_p_value_two_sided = 2 * pnorm(-abs(statistic)),
    dm_p_value_svar_lower_loss = pnorm(statistic)
  )
}

pun_forecast_accuracy <- cbind(
  variable = "pun",
  forecast_metrics(
    recursive_forecasts$actual_pun,
    recursive_forecasts$forecast_pun,
    recursive_forecasts$rw_pun
  ),
  clark_west_test(
    recursive_forecasts$actual_pun,
    recursive_forecasts$forecast_pun,
    recursive_forecasts$rw_pun
  ),
  diebold_mariano_test(
    recursive_forecasts$actual_pun,
    recursive_forecasts$forecast_pun,
    recursive_forecasts$rw_pun
  )
)

pun_forecast_accuracy
```
